---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center"
)
```

# standardlastprofile

<!-- badges: start -->
[![](https://codecov.io/gh/flrd/standardlastprofile/branch/main/graph/badge.svg)](https://codecov.io/gh/flrd/standardlastprofile)
<!-- [![](http://cranlogs.r-pkg.org/badges/last-month/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile) -->
[![R-CMD-check](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package provides data about load profile for electricity from the German Association of Energy and Water Industries (BDEW Bundesverband der Energie- und Wasserwirtschaft e.V.) in a tidy format. 

``` {r, message=FALSE, include = FALSE}
library(standardlastprofile)
library(ggplot2)
library(ggpointless)
library(data.table)
```


```{r readme_example, echo = FALSE, fig.asp = 1.4, fig.retina=2}
# labeler
label_names <- c(
  "saturday"="Saturday",
  'sunday'="Sunday",
  'workday'="Workday"
)

label_fun <- function(x) label_names[[x]]

# plot
ggplot(load_profiles,
       aes(x = as.POSIXct(x = paste(Sys.Date(), timestamp)),
           y = watt,
           color = period)) +
  geom_line() +
  facet_grid(profile ~ day, scales = "free_y",
             labeller = labeller(day = as_labeller(label_names))) +
  scale_x_datetime(NULL, date_labels = "%H") +
  scale_y_continuous(NULL) +
  scale_color_manual(name = NULL,
                     values = c(
                       "winter" = "#961BFA",
                       "summer" = "#FA9529",
                       "transition" = "#0CC792"
                       )) +
  labs(title = "Representative Load Profiles",
       subtitle = "96 x 1/4h-measurements a day [in Watt], based on consumption of 1.000 kWh/a",
       caption = "data: www.bdew.de") +
  theme_minimal() +
  theme(legend.position = "top") +
  theme(strip.text.y.right = element_text(angle = 0)) + 
  theme(
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
  ) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank()) +
  NULL
```

In practice, standardized load profiles are used by an energy suppliers to create an annual consumption forecast for customers (or customer groups) who do not have a modern meter. That is, customers whose electricity consumption is not continuously measured. A load profile is a simplification that does not necessarily correspond to the consumption profile of an individual customer, but is a valid approximation for a larger group of similar customers.

## Installation

You can install the development version of standardlastprofile from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("flrd/standardlastprofile")
```

## About the data
The standardlastprofile package contains one dataset called `load_profiles` used for the plot above. 

``` {r, message=FALSE}
library(standardlastprofile)
data(package = "standardlastprofile")
```

Given a day, a period and a profile the data in `load_profiles` represents a 'characteristic day', e.g. a Sunday in winter versus a workday in summer. There are eleven load profiles for three customer groups. These 96 values per day correspond to the the average quarter-hourly power that is expected if the customer / customer group consumes 1000 kWh/a ("normalized annual consumption"). 

In total the dataset contains 9,504 observations of five variables:

``` {r, message=FALSE}
str(load_profiles)
```

If you have no idea what `H0` etc. stands for, you are not alone. 

- `H0`: households (German: "Haushalte")
- `G0` to `G6`: commerce ("Gewerbe")
- `L0` to `L2`: agriculture ("Landwirtschaft")

Call `get_load_profile_info()` for more information and examples.

``` {r, message=FALSE, echo=TRUE}
get_load_profile_info(language = "EN")$H0
```

### Generate a load profile

Use the function `get_load_profile()` to generate a load profile. 

``` {r, message=FALSE, echo=TRUE, eval=FALSE}
get_load_profile(profile = "G5",
                 start_date = "2023-12-22",
                 end_date = "2023-12-27")
```

The algorithm sets a public holiday to be a Sunday, December 24 and 31 to be a Saturday -- if they are not a Sunday. **Note**: As of now the package supports only public holidays for Germany, which were retrieved from the [nager.Date API](https://github.com/nager/Nager.Date).

``` {r G5_example, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
slp_G5 <- get_load_profile(profile = "G5",
                           start_date = "2023-12-22",
                           end_date = "2023-12-27")

library(ggplot2)
library(ggpointless)
ggplot(slp_G5, aes(start_time, watt)) +
  geom_line(color = "#0CC792") +
  geom_pointless(color = "#0CC792", location = c("first", "last")) + 
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +
  labs(title = "Profile 'G5': bakery with bakehouse",
       subtitle = "1/4h-measurements, based on consumption of 1.000 kWh/a",
       caption = "data: www.bdew.de",
       x = NULL,
       y = "[Watt]") +
  theme_minimal() +
  theme(
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
  ) +
  NULL
```

In contrast to most commercial and agricultural businesses, which have a relatively even consumption and a fairly constant power consumption in the three annual zones, households have a continuously decreasing load from winter to summer and an increasing load towards winter. This is taken into account by applying the dynamization function (4th order polynomial), see [bdew.de](https://www.bdew.de/media/documents/1999_Repraesentative-VDEW-Lastprofile.pdf), page 32/46.

``` {r H0_example, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
slp_H0 <- get_load_profile(profile = "H0",
                           start_date = "2023-01-01",
                           end_date = "2024-01-01") |> 
  as.data.table()

# aggregate by day of year and year
dt_H0 <- slp_H0[, .(start_time = start_time[1], watt = mean(watt)), by = .(day = format(start_time, "%j"), year = format(start_time, "%Y"))]

# plot
ggplot(dt_H0, aes(start_time, watt)) +
  geom_line(color = "#0CC792") +
  geom_pointless(location = c("first", "last"), color = "#0CC792") + 
  labs(title = "Profile 'H0': Households",
       subtitle = "Daily consumption using a dynamic load profile",
       caption = "data: www.bdew.de",
       x = NULL,
       y = "[Watt]") +
  theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  NULL
```

## Source
Data and information about the methodology can be found on website of the BDEW: https://www.bdew.de/energie/standardlastprofile-strom/

## Code of Conduct

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/flrd/standardlastprofile/blob/master/conduct.md). By participating in this project you agree to abide by its terms.
