---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center"
)
```

# standardlastprofile

<!-- badges: start -->
[![](https://codecov.io/gh/flrd/standardlastprofile/branch/main/graph/badge.svg)](https://codecov.io/gh/flrd/standardlastprofile)
<!-- [![](http://cranlogs.r-pkg.org/badges/last-month/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile) -->
[![R-CMD-check](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

This package provides data about representative, standardized load profiles for electricity from the German Association of Energy and Water Industries (BDEW Bundesverband der Energie- und Wasserwirtschaft e.V.) in a tidy format.

``` {r, message=FALSE, include = FALSE}
library(standardlastprofile)
library(ggplot2)
```


```{r small_multiples, echo = FALSE, fig.asp = 1.4, fig.retina=2}
#| fig.alt = "Small multiple line chart of 11 standardized load profiles
#|  published by the German Association of Energy and Water Industries (BDEW 
#|  Bundesverband der Energie- und Wasserwirtschaft e.V.). The lines compare
#|  the consumption for three different periods over a year, and
#|  also compare the consumption between different days of a week."

# labeller
label_names <- c(
  "saturday"="Saturday",
  'sunday'="Sunday",
  'workday'="Workday"
)

label_fun <- function(x) label_names[[x]]

# plot
ggplot(load_profiles,
       aes(x = as.POSIXct(x = paste(Sys.Date(), timestamp)),
           y = watts,
           color = period)) +
  geom_line() +
  facet_grid(profile ~ day, scales = "free_y",
             labeller = labeller(day = as_labeller(label_names))) +
  scale_x_datetime(NULL, date_labels = "%H") +
  scale_y_continuous(NULL) +
  scale_color_manual(name = NULL,
                     values = c(
                       "winter" = "#961BFA",
                       "summer" = "#FA9529",
                       "transition" = "#0CC792"
                       )) +
  labs(title = "Representative Load Profiles",
       subtitle = "96 x 1/4h-measurements a day [in watts], based on consumption of 1.000 kWh/a",
       caption = "data: www.bdew.de") +
  theme_minimal() +
  theme(legend.position = "top") +
  theme(strip.text.y.right = element_text(angle = 0)) + 
  theme(
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank()
  ) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank()) +
  NULL
```

## Installation

You can install the development version of standardlastprofile from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("flrd/standardlastprofile")
```

## About the data
The standardlastprofile package contains one data-set called `load_profiles`. In total there are 9,504 observations of five variables:

- `profile`: one of 11 load profiles
- `period`: one of 'summer', 'winter', 'transition'
- `day`: one of 'saturday', 'sunday', 'workday'
- `timestamp`: format '%H:%M'
- `watts`: electric power

``` {r, message=FALSE}
str(load_profiles)
```
A standardized load profile is a simplification that does not necessarily correspond to the consumption profile of an individual customer, but represents a valid approximation for a larger group of similar customers. These load profiles can be used as a basis for energy utilities to create an annual consumption forecast for their customers. 

For each combination of `profile`, `period` and `day` there are 96 x 1/4h-measurements (in watts). If you have no idea what the profile `H0` stands for, you are not alone. 

- `H0`: households (German: "Haushalte")
- `G0` to `G6`: commerce ("Gewerbe")
- `L0` to `L2`: agriculture ("Landwirtschaft")

There are 11 different `profile`s in total, call `get_load_profile_info()` for more information.

``` {r, message=FALSE, echo=TRUE}
get_load_profile_info(language = "EN")$H0
```

### Generate a load profile

Use the function `get_load_profile()` to generate a load profile. 

``` {r, message=FALSE, echo=TRUE, eval=FALSE}
get_load_profile(profile = "G5",
                 start_date = "2023-12-22",
                 end_date = "2023-12-27")
```

See [`vignette("algorithm-step-by-step", package = "standardlastprofile")`](https://flrd.github.io/standardlastprofile/articles/algorithm-step-by-step.html) for a detailed explanation of the algorithm.
  
``` {r G5_data_readme, message=FALSE, echo=FALSE}
G5 <- get_load_profile(profile = "G5",
                       start_date = "2023-12-22",
                       end_date = "2023-12-27")
```

``` {r G5_plot_readme, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
#| fig.alt = "Line plot of the standardized load profile 'G5' (Bakery
#|  with a bakehouse) from December 22nd to December 27th 2023; values 
#|  are normalized to an annual consumption of 1,000kWh per annum."
ggplot(G5, aes(start_time, watts)) +
  geom_line(color = "#0CC792") +
  scale_x_datetime(
    date_breaks = "1 day",
    date_labels = "%b %d") +
  labs(
    title = "Profile 'G5': bakery with bakehouse",
    subtitle = "1/4h-measurements, based on consumption of 1.000 kWh/a",
    caption = "data: www.bdew.de",
    x = NULL,
    y = "[watts]") +
  theme_minimal() +
  theme(
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
  ) +
  NULL
```

## Source
Data and information about the methodology can be found on website of the BDEW: https://www.bdew.de/energie/standardlastprofile-strom/

## Code of Conduct

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/flrd/standardlastprofile/blob/master/conduct.md). By participating in this project you agree to abide by its terms.
