---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "95%",
  fig.align = "center"
)
```

# standardlastprofile

<!-- badges: start -->
[![R-CMD-check](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/flrd/standardlastprofile/branch/main/graph/badge.svg)](https://app.codecov.io/gh/flrd/standardlastprofile)
[![CRAN version](https://www.r-pkg.org/badges/version/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile)
[![CRAN downloads](http://cranlogs.r-pkg.org/badges/grand-total/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile)
<!-- badges: end -->

Standard load profiles (SLPs) for electricity and gas, published by the German Association of Energy and Water Industries (BDEW Bundesverband der Energie- und Wasserwirtschaft e.V.). SLPs are used by utilities, distribution network operators, and the energy industry to forecast demand for customer groups that are not continuously metered. 

``` {r, message=FALSE, include = FALSE}
library(standardlastprofile)
library(ggplot2)
```

```{r bdew-1999-small_multiples, echo = FALSE, fig.asp = 2.64, fig.retina=2}
#| fig.alt = "Small multiple line chart of 11 electricity standard load profiles
#|  from 1999 published by the German Association of Energy and Water Industries
#|  (BDEW). Lines compare three seasonal periods across different day types."

standardlastprofile:::.slp_plot_1999()
```

## Installation

```{r eval=FALSE}
install.packages("standardlastprofile")
```

## Included features

- `slp_info()` — descriptions for all electricity and gas profile IDs

**Electricity**

- `slp_electricity_profiles` — dataset of BDEW electricity SLPs in tidy format
- `slp_electricity()` — generate a 15-minute profile for any date range

**Gas**

- `slp_gas()` — generate daily gas consumption via the SigLinDe method
- `slp_gas_coefficients()` — retrieve SigLinDe coefficients for gas SLPs
- `slp_gas_kundenwert()` — derive the customer value (German: "Kundenwert") from a reference temperature series
- `slp_gas_siglinde()` — low-level SigLinDe function, can be useful for custom or region-specific SigLinDe coefficients
- `slp_gas_weekday_factors()` — retrieve weekday factors for gas SLPs

## Electricity

The dataset `slp_electricity_profiles` contains 26,784 observations across 5 variables:

- `profile_id`: load profile identifier
- `period`: `"summer"`, `"winter"`, or `"transition"` for 1999 profiles; a lowercase month name for 2025 profiles
- `day`: `"workday"`, `"saturday"`, or `"sunday"`
- `timestamp`: quarter-hour start time in `"%H:%M"` format
- `watts`: average electric power, normalised to 1,000 kWh/a

```{r, message=FALSE}
str(slp_electricity_profiles)
```

### 1999 profiles

Based on an analysis of 1,209 load profiles of low-voltage electricity consumers in Germany[^bdew-1]:

[^bdew-1]: Methodology: <https://www.bdew.de/media/documents/1999_Repraesentative-VDEW-Lastprofile.pdf>

- `H0`: households
- `G0`–`G6`: commercial
- `L0`–`L2`: agriculture

### 2025 profiles

An updated set published by BDEW in 2025, reflecting changes in consumption patterns since the original study. Unlike the 1999 profiles (three seasonal periods), the 2025 profiles provide values for each calendar month:

- `H25`, `G25`, `L25`: updated household, commercial, and agriculture profiles
- `P25`: households with a photovoltaic (PV) system
- `S25`: households with a PV system and battery storage

```{r bdew-2025-small_multiples, echo = FALSE, fig.asp = 1.2, fig.retina = 2}
#| fig.alt = "Small multiple line chart of five electricity standard load profiles
#|  published by BDEW in 2025. Lines are coloured by calendar month and faceted
#|  by profile and day type."

standardlastprofile:::.slp_plot_2025()
```

The chart below compares cumulative energy consumption of the 2025 household profiles against `H0` over a full year. `H25` tracks `H0` closely; `P25` and `S25` flatten from spring through summer as solar generation and storage reduce grid draw.

```{r H0_vs_2025, echo=FALSE, message=FALSE, fig.asp=0.78, fig.retina=2}
#| fig.alt = "Faceted line plot with four panels for H0, H25, P25, and S25.
#|  Each panel shows cumulative energy consumption in kWh over 2026.
#|  A grey reference line shows H0. H25 tracks H0 closely, while P25 and S25
#|  diverge from spring onwards due to photovoltaic generation and battery storage."
dat <- slp_electricity(
  profile_id = c("H0", "H25", "P25", "S25"),
  start_date = "2026-01-01",
  end_date   = "2026-12-31"
)

dat$kWh <- tapply(dat, dat$profile_id, \(x) {
  (cumsum(x$watts) - x$watts[1]) / 4 / 1000
}) |>
  unlist()

ggplot(transform(dat, profile_facet = profile_id), aes(start_time, kWh)) +
  geom_line(
    data = dat[, c("start_time", "kWh", "profile_id")],
    aes(group = profile_id),
    colour = "#CCCCCC"
  ) +
  geom_line(aes(colour = profile_id), linewidth = 1) +
  facet_wrap(vars(profile_facet), ncol = 2) +
  scale_x_datetime(NULL, date_labels = "%b") +
  scale_y_continuous(breaks = c(0, 500, 1000)) +
  scale_colour_manual(
    values = c(
      "H25" = "#961BFA",
      "P25" = "#FA9529",
      "S25" = "#0CC792"
    )
  ) +
  labs(
    title    = "H0 vs. H25, P25, and S25",
    subtitle = "Cumulative energy consumption over the course of one year",
    caption  = "data: www.bdew.de",
    y        = "kWh"
  ) +
  theme_minimal() +
  theme(
    legend.position    = "none",
    strip.text.y.right = element_text(angle = 0),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid = element_line(linetype = "12", lineend = "round", colour = "#FAF6F4")
  )
```

### Generate a profile

`slp_electricity()` returns a data frame with one row per 15-minute interval:

```{r G5_data_readme, message=FALSE}
G5 <- slp_electricity(
  profile_id = "G5",
  start_date = "2023-12-22",
  end_date   = "2023-12-27"
)

head(G5)
```

```{r G5_plot_readme, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
#| fig.alt = "Line plot of the electricity standard load profile 'G5' (bakery
#|  with a bakehouse) from December 22nd to 27th 2023, normalised to 1,000 kWh/a."
ggplot(G5, aes(start_time, watts)) +
  geom_line(colour = "#0CC792") +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%b %d") +
  scale_y_continuous(NULL, labels = \(x) paste(x, "W")) +
  labs(
    title    = "'G5': bakery with bakehouse",
    subtitle = "1/4h measurements, based on consumption of 1,000 kWh/a",
    caption  = "data: www.bdew.de",
    x        = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid = element_line(linetype = "12", lineend = "round", colour = "#FAF6F4")
  )
```

### Public holidays

Both `slp_electricity()` and `slp_gas()` use the same holiday logic: nine
nationwide German public holidays are treated as Sundays by default (New
Year's, Good Friday, Easter Monday, Labour Day, Ascension Day, Whit Monday,
German Unity Day, Christmas Day, Boxing Day). State-level holidays are not
included because they vary by state and can change. Use the `holidays`
argument in either function to supply your own dates — the built-in data
are then ignored entirely:

```{r holidays-berlin, eval=FALSE}
library(httr2)

resp <- request("https://date.nager.at") |>
  req_url_path_append("api", "v3", "PublicHolidays", "2027", "DE") |>
  req_perform() |>
  resp_body_json()

# Berlin observes International Women's Day (8 March) in addition to all
# nationwide holidays; global == TRUE means observed in all states
is_berlin <- \(x) isTRUE(x$global) || "DE-BE" %in% unlist(x$counties)

holidays_berlin_2027 <- as.Date(
  vapply(Filter(is_berlin, resp), \(x) x$date, character(1))
)

# electricity
slp_electricity("H0", "2027-01-01", "2027-12-31",
                holidays = holidays_berlin_2027)

# gas — same holidays argument, same semantics
slp_gas("HEF", dates_2027, temps_2027, kundenwert = kw,
        holidays = holidays_berlin_2027)
```

## Gas

`slp_gas()` implements the [BDEW/VKU/GEODE synthetic procedure (SigLinDe method)](https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf)
for daily gas consumption. It takes daily mean temperatures and a customer
value (`kundenwert`, kWh/day), and supports all 15 gas profile IDs.

Take a single-family home (profile `HEF`) in Düsseldorf with a `kundenwert` of
55.1 kWh/day. First grab the daily mean temperatures for the period of
interest — here the 2025/26 heating season — from the DWD open-data archive
via [`rdwd`](https://cran.r-project.org/package=rdwd) (no API key required;
`TMK` is the daily mean temperature in °C):

```{r slp_gas_fetch, eval=FALSE}
library(rdwd)

dates <- seq.Date(as.Date("2025-10-01"), as.Date("2026-04-30"), by = "day")

# Düsseldorf = DWD station 1078. The "recent" file is a rolling ~550-day
# window, so combine it with "historical" to cover any period; force = TRUE
# avoids reusing a stale cached download.
link  <- selectDWD("Duesseldorf", res = "daily", var = "kl",
                   per = c("historical", "recent"))
raw   <- do.call(rbind, readDWD(dataDWD(link, read = FALSE, force = TRUE),
                                varnames = FALSE))
temps <- raw$TMK[match(dates, as.Date(raw$MESS_DATUM))]
stopifnot(!anyNA(temps))   # fail loudly if the period isn't fully covered
```

```{r slp_gas_setup, include=FALSE}
# Real DWD values for the fetch above, embedded so the README builds offline.
dates <- seq.Date(as.Date("2025-10-01"), as.Date("2026-04-30"), by = "day")
temps <- c(
  10.2, 11.6, 11.7, 12.5, 11.6, 12.0, 14.2, 14.4, 13.6, 14.8, 14.1, 14.0,
  13.8, 12.1, 12.4, 11.8, 12.1, 9.8, 10.6, 14.6, 13.4, 12.8, 12.6, 9.3,
  9.3, 7.7, 9.0, 10.4, 12.2, 10.3, 10.6, 12.2, 10.8, 9.9, 13.7, 12.6,
  11.6, 10.3, 8.3, 10.3, 7.5, 10.4, 11.9, 14.3, 15.1, 12.4, 8.1, 3.5,
  3.6, 4.6, 2.3, 0.9, 0.6, 1.2, 5.3, 5.5, 3.7, 3.4, 7.2, 8.9, 7.1, 5.1,
  5.8, 7.3, 5.3, 5.3, 7.6, 10.5, 12.5, 12.3, 11.6, 8.5, 8.5, 7.6, 6.1,
  6.7, 8.3, 7.3, 9.8, 10.4, 5.4, 8.9, 3.5, 4.8, 0.3, -2.9, -1.5, 0.2,
  1.2, 0.0, 0.2, 0.2, 3.3, 1.4, 0.5, 0.3, -1.6, -1.1, -1.5, 2.0, 3.8,
  -1.8, -3.3, 3.5, 8.8, 8.4, 9.8, 9.8, 8.2, 2.7, 1.7, 2.4, 4.5, 3.0, 5.5,
  3.6, 0.2, 0.0, 2.8, 3.3, 1.1, 2.5, 5.6, 5.3, 5.0, 2.3, 5.9, 4.5, 5.0,
  8.3, 4.9, 4.1, 6.7, 9.6, 8.7, 4.7, 0.5, -0.4, 4.3, 4.1, 3.1, 2.7, 3.8,
  9.4, 11.6, 9.8, 10.4, 12.4, 13.0, 13.7, 10.9, 8.7, 10.9, 11.2, 9.4,
  11.5, 12.3, 10.3, 10.0, 11.6, 11.1, 9.4, 8.8, 8.1, 3.7, 6.0, 7.3, 9.4,
  10.8, 8.9, 8.5, 8.6, 8.8, 9.1, 11.4, 8.4, 3.9, 4.9, 6.3, 7.0, 7.0, 5.5,
  6.5, 7.1, 8.9, 12.8, 12.1, 8.4, 10.3, 12.1, 15.5, 9.8, 11.9, 8.8, 7.7,
  9.3, 12.9, 14.3, 13.7, 13.3, 10.6, 8.2, 9.9, 11.0, 10.8, 9.9, 9.5,
  10.1, 11.6, 13.8, 13.1, 13.5
)
```

Then pass `dates` and `temps` to `slp_gas()` together with the `kundenwert`:

```{r slp_gas_readme, message=FALSE}
HEF <- slp_gas("HEF", dates, temps, kundenwert = 55.1)
head(HEF)

# total gas consumption over the heating season (1 Oct 2025 – 30 Apr 2026)
sum(HEF$kwh)
```

> The Kundenwert of 55.1 kWh/day is itself derived once, from the customer's
> annual consumption and a reference temperature series. See the
> [gas article](https://flrd.github.io/standardlastprofile/articles/slp-gas.html)
> for that step and the full method.

```{r slp_gas_readme_plot, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
#| fig.alt = "Line chart of daily gas consumption in kilowatt-hours for a
#|  single-family home in Düsseldorf across the 2025/26 heating season. Demand
#|  peaks in the cold winter months and is lowest in the mild shoulder months of
#|  October and April."
ggplot(HEF, aes(date, kwh)) +
  geom_line(colour = "#0CC792") +
  labs(
    title = "Single family home in Düsseldorf",
    subtitle = "Heating period from October '25 to April '26",
    caption  = "data: www.bdew.de",
    x        = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid = element_line(linetype = "12", lineend = "round", colour = "#FAF6F4")
  )
```


Let's assume that same customer (i.e. fixed `kundenwert`) would live in another
place with a different climate. The chart below uses temperature data for the same
heating season (October '25 to April '26) to compare their daily gas consumption in:
- Chemnitz
- Freiburg im Breisgau
- Hamburg
against Düsseldorf. 

Points above the 45° line mean that customer would have consumed more gas than
in Düsseldorf due to different climate and temperatures. We see that in that
winter all three cities ran colder than Düsseldorf, so every cloud sits above
the line — most for Chemnitz (~35% more gas over the season), least for 
Freiburg (~11% more), with Hamburg in between (~25% more):

```{r slp_gas_cities, echo=FALSE, fig.asp=2.4, fig.retina=2, message=FALSE}
#| fig.alt = "Faceted scatterplot grid: columns are Chemnitz, Freiburg, Hamburg;
#|  rows are months October to April. Each point is a day; the x-axis is daily
#|  gas consumption in Düsseldorf, the y-axis in the comparison city, with a
#|  45-degree reference line. All three cities' point clouds sit above the line
#|  in winter, most clearly for Chemnitz."
standardlastprofile:::.slp_plot_gas_cities()
```

For a detailed walkthrough of the SigLinDe parameters and the full climate zone comparison, see the [gas articles](https://flrd.github.io/standardlastprofile/articles/index.html) on the package website.

## Sources

- Electricity SLPs: <https://www.bdew.de/energie/standardlastprofile-strom/>
- Gas SLPs: <https://www.bdew.de/energie/standardlastprofile-gas/>

## Code of Conduct

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/flrd/standardlastprofile/blob/main/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
