---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "90%",
  fig.align = "center"
)
```

# standardlastprofile

<!-- badges: start -->
<!-- [![](https://www.r-pkg.org/badges/version/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile) -->
<!-- [![](http://cranlogs.r-pkg.org/badges/last-month/standardlastprofile)](https://cran.r-project.org/package=standardlastprofile) -->
[![R-CMD-check](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/flrd/standardlastprofile/actions/workflows/R-CMD-check.yaml)
[![](https://codecov.io/gh/flrd/standardlastprofile/branch/main/graph/badge.svg)](https://app.codecov.io/gh/flrd/standardlastprofile)
<!-- badges: end -->

This package provides data on representative, standard load profiles for electricity from the German Association of Energy and Water Industries (BDEW Bundesverband der Energie- und Wasserwirtschaft e.V.) in a tidy format.

``` {r, message=FALSE, include = FALSE}
library(standardlastprofile)
library(ggplot2)
```


```{r small_multiples, echo = FALSE, fig.asp = 1.4, fig.retina=2}
#| fig.alt = "Small multiple line chart of 11 standard load profiles
#|  published by the German Association of Energy and Water Industries (BDEW 
#|  Bundesverband der Energie- und Wasserwirtschaft e.V.). The lines compare
#|  the consumption for three different periods over a year, and
#|  also compare the consumption between different days of a week."

# labeller
label_names <- c(
  "saturday" = "Saturday",
  "sunday" = "Sunday",
  "workday" = "Workday"
)

label_fun <- function(x) label_names[[x]]

# plot
ggplot(slp,
       aes(x = as.POSIXct(x = paste(Sys.Date(), timestamp)),
           y = watts,
           color = period)) +
  geom_line() +
  facet_grid(profile_id ~ day, scales = "free_y",
             labeller = labeller(day = as_labeller(label_names))) +
  scale_x_datetime(NULL, date_labels = "%H") +
  scale_y_continuous(NULL) +
  scale_color_manual(name = NULL,
                     values = c(
                       "winter" = "#961BFA",
                       "summer" = "#FA9529",
                       "transition" = "#0CC792"
                       )) +
  labs(title = "Standard Load Profiles",
       subtitle = "96 x 1/4h measurements a day [in watts], based on consumption of 1,000 kWh/a",
       caption = "data: www.bdew.de") +
  theme_minimal() +
  theme(legend.position = "top") +
  theme(strip.text.y.right = element_text(angle = 0)) + 
  theme(
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank()
  ) +
  theme(axis.text.x = element_blank()) +
  theme(axis.text.y = element_blank()) +
  NULL
```

## Installation

<!-- You can install the released version of standardlastprofile from CRAN with: -->
<!-- ```{r eval=FALSE} -->
<!-- install.packages("standardlastprofile") -->
<!-- ``` -->

To install the development version of standardlastprofile from 
[GitHub](https://github.com/), call:

``` {r eval=FALSE}
# install.packages("devtools")
devtools::install_github("flrd/standardlastprofile")
```

## What will you get

The following dataset and functions are implemented:

- `slp` -- Dataset for BDEW standard load profiles in electricity
- `slp_generate()` --	Interface to generate a standard load profile for a user-defined period of time
- `slp_info()` -- Function to retrieve details of BDEW standard load profiles

## About the data
The standardlastprofile package contains one dataset called `slp`. The data was published in 1999. They are based on an analysis of 1,209 load profiles of low-voltage consumers in Germany.[^bdew-2] In total there are 9,504 observations of 5 variables:

  [^bdew-2]: More information on the data and methodology can be found [here](https://www.bdew.de/media/documents/1999_Repraesentative-VDEW-Lastprofile.pdf).

- `profile_id`: identifier of standard load profile
- `period`: one of "summer", "winter", "transition"
- `day`: one of "saturday", "sunday", "workday"
- `timestamp`: format "%H:%M"
- `watts`: electric power

``` {r, message=FALSE}
str(slp)
```

In the context of the German electricity market, the term 'Standard Load Profile' denotes a representative pattern of electricity consumption over a specific period. These profiles portray anticipated electricity consumption for diverse customer groups, like households or businesses. While not an exact match for an individual customer's consumption profile, they serve as a valid approximation for larger groups of similar customers.

For each unique combination of `profile_id`, `period` and `day` there are 
96 x 1/4 hour measurements in watts. If `H0` keeps you puzzled,
you are not alone:

- `H0`: households (German: "Haushalte")
- `G0` to `G6`: commerce ("Gewerbe")
- `L0` to `L2`: agriculture ("Landwirtschaft")

For more details, call the `slp_info()` function.

``` {r, message=FALSE, echo=TRUE}
slp_info(profile_id = "H0", language = "EN")
```

### Generate a standard load profile

Generate a standard load profile for a user-defined period of time using the `slp_generate()` function:

``` {r, G5_data_readme, message=FALSE, echo=TRUE}
G5 <- slp_generate(
  profile_id = "G5",
  start_date = "2023-12-22",
  end_date = "2023-12-27"
  )
head(G5)
```


``` {r G5_plot_readme, message=FALSE, echo=FALSE, fig.retina=2, fig.asp=0.5}
#| fig.alt = "Line plot of the standard load profile 'G5' (i.e. Bakery
#|  with a bakehouse) based on data from the German Association of Energy
#|  and Water Industries (BDEW Bundesverband der Energie- und Wasserwirtschaft
#|  e.V.) from December 22nd to December 27th 2023; values 
#|  are normalized to an annual consumption of 1,000 kWh."
ggplot(G5, aes(start_time, watts)) +
  geom_line(color = "#0CC792") +
  scale_x_datetime(
    date_breaks = "1 day",
    date_labels = "%b %d") +
  labs(
    title = "'G5': bakery with bakehouse",
    subtitle = "1/4h-measurements, based on consumption of 1,000 kWh/a",
    caption = "data: www.bdew.de",
    x = NULL,
    y = "[watts]") +
  theme_minimal() +
  theme(
      panel.grid.minor.x = element_blank(),
      panel.grid.minor.y = element_blank()
  ) +
  NULL
```

See [`vignette("algorithm-step-by-step", package = "standardlastprofile")`](https://flrd.github.io/standardlastprofile/articles/algorithm-step-by-step.html) for a detailed explanation of the algorithm.

## Source
Data and information about the methodology can be found on website of the BDEW: https://www.bdew.de/energie/standardlastprofile-strom/

## Code of Conduct

Please note that this project is released with a [Contributor Code of Conduct](https://github.com/flrd/standardlastprofile/blob/main/CODE_OF_CONDUCT.md). 
By participating in this project you agree to abide by its terms.
