---
title: "Deriving the Kundenwert from a reference temperature series"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Deriving the Kundenwert from a reference temperature series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  dpi       = 300,
  fig.asp   = 0.618,
  fig.width = 6,
  fig.align = "center",
  out.width = "90%"
)
library(standardlastprofile)
library(ggplot2)
has_plotly  <- requireNamespace("plotly",  quietly = TRUE)
has_leaflet <- requireNamespace("leaflet", quietly = TRUE)
```

When `kundenwert = NULL` (the default), `slp_gas()` derives $KW$
automatically from `annual_consumption` and the supplied temperature series.
You can also supply a fixed `kundenwert` directly, which is useful when you
already know the customer's daily baseline and want to bypass the annual
normalisation.

The daily gas quantity $Q(D)$ is computed as $Q(D) = KW \cdot h(\vartheta_D)
\cdot F_{WT}$, where $KW$ — the *Kundenwert* (kWh/day) — scales the
dimensionless profile values to match a customer's actual annual consumption.
When you pass daily temperatures for a short period (e.g. a single month) to
`slp_gas()` without supplying `kundenwert`, the function must derive
$KW$ from that short series alone. Because the denominator
$\sum_D h(\vartheta_D) \cdot F_{WT,D}$ then reflects only the temperatures
you happened to observe, $KW$ absorbs the missing seasonal variation — in the
extreme case of a single day, the $h$ values cancel and `kwh` equals
`annual_consumption` exactly, regardless of temperature.

The remedy is a **two-step workflow**:

1. Compute $KW$ once from a **full reference year** of temperatures using
   `slp_gas_kundenwert()`.
2. Pass the resulting `kundenwert` to `slp_gas()` for any
   sub-period you care about.

## Built-in reference stations

The package ships with long-term mean daily temperatures (WMO climate normal
period 1991--2020) for ten representative DWD weather stations covering the
major German climate regions. Pick the station closest to your customer's
location and pass it via the `station` argument of `slp_gas_kundenwert()`.

```{r station_map, echo=FALSE}
if (has_leaflet) {
  stn <- data.frame(
    station    = c("Oberstdorf", "Potsdam", "Hamburg", "Freiburg",
                   "Chemnitz", "Duesseldorf", "Erfurt",
                   "Frankfurt", "Nuernberg", "Regensburg"),
    region     = c("Alpenrand / Allgaeu", "Kontinentales Binnenland",
                   "Maritime Nordseekueste", "Oberrheingraben",
                   "Erzgebirge / Mittelgebirge", "Niederrhein",
                   "Thueringer Becken",
                   "Rhein-Main-Gebiet", "Mittelfranken",
                   "Oberpfalz / Donau"),
    dwd_id     = c(3730L, 3987L, 1975L, 1443L, 853L, 1078L, 1270L,
                   1420L, 3668L, 4104L),
    elevation  = c(806L, 81L, 11L, 237L, 416L, 37L, 316L,
                   112L, 314L, 365L),
    lat        = c(47.4075, 52.3813, 53.5503, 47.9990, 50.7913, 51.2959,
                   50.9787, 50.0259, 49.5030, 49.0420),
    lon        = c(10.2783, 13.0622, 9.9872, 7.8523, 12.8726, 6.7681,
                   10.9632, 8.5958, 11.0549, 12.1018),
    mean_temp  = c(7.0, 9.8, 9.8, 11.6, 9.0, 11.1, 9.0,
                   11.1, 9.8, 9.4),
    stringsAsFactors = FALSE
  )

  leaflet::leaflet(stn, width = "100%", height = 420) |>
    leaflet::addTiles() |>
    leaflet::addCircleMarkers(
      lng = ~lon, lat = ~lat, radius = 7,
      color = "#0CC792", fillColor = "#0CC792", fillOpacity = 0.8,
      stroke = TRUE, weight = 2,
      label = ~paste0(
        station, " \u2014 DWD ", dwd_id,
        " | ", elevation, " m",
        " | \u00d8 ", format(mean_temp, nsmall = 1), " \u00b0C"
      ),
      labelOptions = leaflet::labelOptions(
        style = list("font-size" = "13px", "padding" = "4px 8px"),
        direction = "top", offset = c(0, -10)
      )
    )
}
```

## Explore the SigLinDe profile functions

The dimensionless profile value $h(\vartheta)$ describes how gas demand changes
with temperature. Select profiles and switch between variants to see how
the curves differ. Click a legend entry to show or hide a profile; double-click
to isolate it.

```{r h_curve_precompute, include=FALSE}
theta <- seq(-15, 30, by = 0.5)
all_profiles <- c("HEF", "HMF", "HKO", "GKO", "GHA", "GMK", "GBD", "GBH",
                  "GWA", "GGA", "GBA", "GGB", "GPD", "GMF", "GHD")
key_profiles <- c("HEF", "HMF", "HKO", "GKO", "GHA")

h_data <- list()
for (v in c("34", "33")) {
  for (pid in all_profiles) {
    p <- standardlastprofile:::.gas_profile_params[[v]][[pid]]
    h_data[[paste(v, pid, sep = "|")]] <- slp_gas_siglinde(
      theta, p$A, p$B, p$C, p$D, p$theta0, p$mH, p$bH, p$mW, p$bW
    )
  }
}
```

```{r h_curve_interactive, echo=FALSE, fig.cap="SigLinDe profile functions h(\u03b8) across temperature. Toggle variant with the buttons; click legend entries to show/hide profiles."}
if (has_plotly) {
  n <- length(all_profiles)

  # Colour palette: viridis-like but more distinct
  pal <- c(
    "#0CC792", "#1B9E77", "#D95F02", "#7570B3", "#E7298A",
    "#66A61E", "#E6AB02", "#A6761D", "#666666", "#1F78B4",
    "#B2DF8A", "#FB9A99", "#FDBF6F", "#CAB2D6", "#FF7F00"
  )

  fig <- plotly::plot_ly()

  # -- Variant 34 traces (initially visible) --
  for (i in seq_along(all_profiles)) {
    pid <- all_profiles[i]
    vis <- if (pid %in% key_profiles) TRUE else "legendonly"
    fig <- plotly::add_trace(
      fig,
      x = theta, y = h_data[[paste("34", pid, sep = "|")]],
      type = "scatter", mode = "lines",
      name = pid, legendgroup = pid,
      visible = vis, showlegend = TRUE,
      line = list(width = 2.5, color = pal[i])
    )
  }

  # -- Variant 33 traces (initially hidden) --
  for (i in seq_along(all_profiles)) {
    pid <- all_profiles[i]
    fig <- plotly::add_trace(
      fig,
      x = theta, y = h_data[[paste("33", pid, sep = "|")]],
      type = "scatter", mode = "lines",
      name = pid, legendgroup = pid,
      visible = FALSE, showlegend = FALSE,
      line = list(width = 2.5, color = pal[i])
    )
  }

  # Visibility arrays for buttons
  vis_34 <- c(
    ifelse(all_profiles %in% key_profiles, TRUE, "legendonly"),
    rep(FALSE, n)
  )
  vis_33 <- c(
    rep(FALSE, n),
    ifelse(all_profiles %in% key_profiles, TRUE, "legendonly")
  )
  vis_both <- c(
    ifelse(all_profiles %in% key_profiles, TRUE, "legendonly"),
    ifelse(all_profiles %in% key_profiles, TRUE, "legendonly")
  )
  # For "both" mode, show legend only for v34 traces
  sl_both <- c(rep(TRUE, n), rep(FALSE, n))

  fig <- plotly::layout(
    fig,
    xaxis = list(
      title = "Temperature (\u00b0C)",
      zeroline = FALSE, gridcolor = "#f0eded"
    ),
    yaxis = list(
      title = "h(\u03b8)",
      zeroline = FALSE, gridcolor = "#f0eded"
    ),
    legend = list(orientation = "h", y = -0.15, x = 0.5, xanchor = "center"),
    updatemenus = list(
      list(
        type = "buttons",
        direction = "right",
        x = 0, y = 1.12,
        xanchor = "left",
        buttons = list(
          list(
            label = "Variant 34 (57% linear)",
            method = "update",
            args = list(
              list(visible = vis_34, showlegend = c(rep(TRUE, n), rep(FALSE, n))),
              list(title = "SigLinDe profile functions \u2014 Auspr\u00e4gung 34")
            )
          ),
          list(
            label = "Variant 33 (45% linear)",
            method = "update",
            args = list(
              list(visible = vis_33, showlegend = c(rep(FALSE, n), rep(TRUE, n))),
              list(title = "SigLinDe profile functions \u2014 Auspr\u00e4gung 33")
            )
          ),
          list(
            label = "Compare both",
            method = "update",
            args = list(
              list(visible = vis_both, showlegend = sl_both),
              list(title = "SigLinDe \u2014 solid: Var. 34, same colour: Var. 33")
            )
          )
        )
      )
    ),
    title = list(
      text = "SigLinDe profile functions \u2014 Auspr\u00e4gung 34",
      x = 0, xanchor = "left"
    ),
    margin = list(t = 60)
  )

  # In "both" mode, dash variant 33 lines
  for (i in seq_along(all_profiles)) {
    fig$x$attrs[[n + i]]$line$dash <- "dot"
  }

  fig
}
```

## Interactive Kundenwert calculator

The Kundenwert is $KW = JVP \;/\; M_{SLP}$ where $JVP$ is the annual
consumption forecast and $M_{SLP} = \sum_D h(\vartheta_D) \cdot F_{WT,D}$
integrates the profile function over a full reference year.  Because $M_{SLP}$
depends on local temperatures, the **same annual consumption produces different
Kundenwerte in different climate regions** (see Leitfaden p. 37: Niederrhein
vs. Allgäu).

The Leitfaden recommends computing $M_{SLP}$ from climate-typical
temperatures — either a multi-year mean (e.g. 3–10 years) or a representative
planning year.[^mslp_basis] The built-in reference temperatures in this package
follow that guidance: they are WMO climate normals (1991–2020 mean) for each
station, providing a 30-year average that smooths out single-year anomalies.

[^mslp_basis]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Abschnitt 3.6.3, p. 38.

Select a profile, variant, and two stations to compare. Enter an annual
consumption (JVP) to see the resulting Kundenwerte side by side.

```{r mslp_precompute, include=FALSE}
# Pre-compute M_SLP for every profile x variant x station combination
all_stations <- unique(slp_gas_normtemperatur$station)
variants     <- c("34", "33")

mslp_list <- list()
for (v in variants) {
  for (s in all_stations) {
    ref <- slp_gas_normtemperatur[
      slp_gas_normtemperatur$station == s,
    ]
    for (pid in all_profiles) {
      kw <- slp_gas_kundenwert(pid, dates = ref$date, temperatures = ref$temp_c_mean,
                            annual_consumption = 1000, variant = v)
      mslp_list[[paste(v, s, pid, sep = "|")]] <- round(1000 / as.numeric(kw), 4)
    }
  }
}

# Build JSON string (avoid jsonlite dependency)
json_entries <- vapply(names(mslp_list), function(key) {
  sprintf('"%s":%.4f', key, mslp_list[[key]])
}, character(1))
mslp_json <- paste0("{", paste(json_entries, collapse = ","), "}")

# Profile descriptions for the dropdown
gas_desc <- c(
  HEF = "HEF \u2014 Single-family home",
  HMF = "HMF \u2014 Multi-family home",
  HKO = "HKO \u2014 Cooking and hot water only",
  GKO = "GKO \u2014 Small commercial",
  GHA = "GHA \u2014 Trade and commerce",
  GMK = "GMK \u2014 Metal and automotive",
  GBD = "GBD \u2014 Services",
  GBH = "GBH \u2014 Accommodation",
  GWA = "GWA \u2014 Laundries",
  GGA = "GGA \u2014 Gastronomy",
  GBA = "GBA \u2014 Bakeries",
  GGB = "GGB \u2014 Mixed commercial",
  GPD = "GPD \u2014 Paper and printing",
  GMF = "GMF \u2014 Large multi-family / mixed use",
  GHD = "GHD \u2014 Trade, commerce and services aggregate"
)
```

```{r kw_calculator, echo=FALSE, results='asis'}
# Build select options
profile_options <- paste0(
  vapply(all_profiles, function(pid) {
    sel <- if (pid == "HEF") " selected" else ""
    sprintf('        <option value="%s"%s>%s</option>', pid, sel, gas_desc[pid])
  }, character(1)),
  collapse = "\n"
)

station_options_a <- paste0(
  vapply(all_stations, function(s) {
    sel <- if (s == "Duesseldorf") " selected" else ""
    sprintf('          <option value="%s"%s>%s</option>', s, sel, s)
  }, character(1)),
  collapse = "\n"
)

station_options_b <- paste0(
  vapply(all_stations, function(s) {
    sel <- if (s == "Oberstdorf") " selected" else ""
    sprintf('          <option value="%s"%s>%s</option>', s, sel, s)
  }, character(1)),
  collapse = "\n"
)

html_template <- '
<style>
.kw-calc { font-family: -apple-system, BlinkMacSystemFont, "Segoe UI", Roboto, sans-serif; max-width: 720px; margin: 1.5em auto; }
.kw-controls { display: flex; flex-wrap: wrap; gap: 12px; margin-bottom: 16px; align-items: end; }
.kw-ctrl { display: flex; flex-direction: column; gap: 4px; }
.kw-ctrl label { font-size: 0.82em; font-weight: 600; color: #555; }
.kw-ctrl select, .kw-ctrl input[type="number"] { padding: 6px 10px; border: 1px solid #ccc; border-radius: 6px; font-size: 0.95em; background: #fff; }
.kw-ctrl input[type="number"] { width: 130px; }
.kw-variant-group { display: flex; gap: 12px; padding-top: 2px; }
.kw-variant-group label { font-weight: 400; font-size: 0.95em; cursor: pointer; }
.kw-cards { display: grid; grid-template-columns: 1fr 1fr; gap: 16px; margin-bottom: 12px; }
.kw-card { border: 1px solid #ddd; border-radius: 10px; padding: 16px; background: #fafafa; }
.kw-card select { width: 100%%; padding: 6px 10px; border: 1px solid #ccc; border-radius: 6px; font-size: 0.95em; margin-bottom: 12px; background: #fff; }
.kw-card-label { font-size: 0.78em; text-transform: uppercase; letter-spacing: 0.05em; color: #888; margin-bottom: 6px; font-weight: 600; }
.kw-metric { display: flex; justify-content: space-between; padding: 6px 0; border-bottom: 1px solid #eee; font-size: 0.95em; }
.kw-metric:last-child { border-bottom: none; }
.kw-metric-label { color: #666; }
.kw-metric-value { font-weight: 600; font-variant-numeric: tabular-nums; }
.kw-explanation { font-size: 0.9em; color: #555; line-height: 1.5; padding: 12px 14px; background: #f5f9f7; border-left: 3px solid #0CC792; border-radius: 0 6px 6px 0; }
</style>

<div class="kw-calc">
<div class="kw-controls">
<div class="kw-ctrl">
<label>Profile</label>
<select id="kw-profile">
%s
</select>
</div>
<div class="kw-ctrl">
<label>Variant</label>
<div class="kw-variant-group">
<label><input type="radio" name="kw-variant" value="34" checked> 34 (57%% linear)</label>
<label><input type="radio" name="kw-variant" value="33"> 33 (45%% linear)</label>
</div>
</div>
<div class="kw-ctrl">
<label>Annual consumption (kWh/a)</label>
<input type="number" id="kw-jvp" value="15000" min="0" step="500">
</div>
</div>
<div class="kw-cards">
<div class="kw-card">
<div class="kw-card-label">Station A</div>
<select id="kw-station-a">
%s
</select>
<div class="kw-metric">
<span class="kw-metric-label">M<sub>SLP</sub></span>
<span class="kw-metric-value" id="kw-mslp-a">&mdash;</span>
</div>
<div class="kw-metric">
<span class="kw-metric-label">Kundenwert</span>
<span class="kw-metric-value" id="kw-value-a">&mdash;</span>
</div>
</div>
<div class="kw-card">
<div class="kw-card-label">Station B</div>
<select id="kw-station-b">
%s
</select>
<div class="kw-metric">
<span class="kw-metric-label">M<sub>SLP</sub></span>
<span class="kw-metric-value" id="kw-mslp-b">&mdash;</span>
</div>
<div class="kw-metric">
<span class="kw-metric-label">Kundenwert</span>
<span class="kw-metric-value" id="kw-value-b">&mdash;</span>
</div>
</div>
</div>
<div class="kw-explanation" id="kw-explanation"></div>
</div>

<script>
(function() {
  var mslpData = %s;
  function fmt(x, d) {
    return x.toLocaleString("en", {minimumFractionDigits: d, maximumFractionDigits: d});
  }
  function update() {
    var profile = document.getElementById("kw-profile").value;
    var variant = document.querySelector("input[name=\\"kw-variant\\"]:checked").value;
    var jvp     = parseFloat(document.getElementById("kw-jvp").value);
    var stA     = document.getElementById("kw-station-a").value;
    var stB     = document.getElementById("kw-station-b").value;
    if (isNaN(jvp) || jvp <= 0) {
      document.getElementById("kw-explanation").innerHTML =
        "Please enter a positive annual consumption.";
      return;
    }
    var keyA = variant + "|" + stA + "|" + profile;
    var keyB = variant + "|" + stB + "|" + profile;
    var mA   = mslpData[keyA];
    var mB   = mslpData[keyB];
    if (!mA || !mB) return;
    var kwA = jvp / mA;
    var kwB = jvp / mB;
    document.getElementById("kw-mslp-a").textContent  = fmt(mA, 1);
    document.getElementById("kw-mslp-b").textContent  = fmt(mB, 1);
    document.getElementById("kw-value-a").textContent  = fmt(kwA, 2) + " kWh/day";
    document.getElementById("kw-value-b").textContent  = fmt(kwB, 2) + " kWh/day";
    var el = document.getElementById("kw-explanation");
    if (stA === stB) {
      el.innerHTML =
        "An annual consumption of <strong>" + fmt(jvp, 0) + " kWh/a</strong>" +
        " in " + stA + " yields KW = <strong>" + fmt(kwA, 2) +
        "</strong> kWh/day (M<sub>SLP</sub> = " + fmt(mA, 1) +
        "). Select two different stations to compare how climate affects the Kundenwert.";
      return;
    }
    var pctDiff = (Math.abs(kwA - kwB) / Math.min(kwA, kwB) * 100).toFixed(1);
    var colder = mA > mB ? stA : stB;
    var milder = mA > mB ? stB : stA;
    el.innerHTML =
      "The same annual consumption of <strong>" + fmt(jvp, 0) + " kWh/a</strong>" +
      " yields KW = <strong>" + fmt(kwA, 2) + "</strong> kWh/day in " + stA +
      " vs. <strong>" + fmt(kwB, 2) + "</strong> kWh/day in " + stB +
      " (a difference of " + pctDiff + "%%). " +
      colder + " has a larger M<sub>SLP</sub> (" +
      fmt(Math.max(mA, mB), 1) + " vs. " + fmt(Math.min(mA, mB), 1) +
      "), reflecting its colder climate: the annual consumption is spread " +
      "across more heating-degree days, so each day receives less gas.";
  }
  ["kw-profile", "kw-station-a", "kw-station-b", "kw-jvp"].forEach(function(id) {
    document.getElementById(id).addEventListener("input", update);
  });
  document.querySelectorAll("input[name=\\"kw-variant\\"]").forEach(function(el) {
    el.addEventListener("change", update);
  });
  update();
})();
</script>
'
cat(sprintf(html_template, profile_options, station_options_a, station_options_b, mslp_json))
```

## Step-by-step example

### Computing the Kundenwert in R

```{r compute_kw}
kw_hef <- slp_gas_kundenwert(
  profile_id         = "HEF",
  station            = "Hamburg",
  annual_consumption = 15000       # kWh/a — a typical single-family home
)
kw_hef
```

Verify that applying KW over the full reference year reproduces the annual
consumption:

```{r verify_kw}
ref_hh <- slp_gas_normtemperatur[
  slp_gas_normtemperatur$station == "Hamburg",
]
q_ref <- slp_gas("HEF", ref_hh$date, ref_hh$temp_c_mean, kundenwert = kw_hef)
sum(q_ref$kwh)   # should equal 15 000
```

### Applying the Kundenwert to a shorter period

With `kundenwert` fixed, `slp_gas()` can be called for any sub-period
without any short-series scaling artefact:

```{r apply_kw}
# Synthetic temperatures for January–March 2025
dates_q1 <- seq.Date(as.Date("2025-01-01"), as.Date("2025-03-31"), by = "day")
doy_q1   <- as.integer(format(dates_q1, "%j"))
set.seed(2025)
temps_q1 <- 10 - 11 * cos(2 * pi * (doy_q1 - 15) / 365) + rnorm(90, sd = 2.5)

q1_hef <- slp_gas("HEF", dates_q1, temps_q1, kundenwert = kw_hef)
head(q1_hef)
```

```{r plot_q1, echo=FALSE, fig.retina=2}
ggplot(q1_hef, aes(date, kwh)) +
  geom_col(fill = "#0CC792", width = 0.8) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%d %b") +
  scale_y_continuous("Daily gas consumption (kWh)") +
  labs(
    title    = "HEF \u2014 daily gas consumption, Q1 2025",
    subtitle = "Annual consumption 15,000 kWh/a \u00b7 Kundenwert from Hamburg reference station",
    x        = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid = element_line(linetype = "12", lineend = "round",
                              colour = "#FAF6F4")
  )
```

Compare this with the naive approach that derives KW from the 90-day series
alone:

```{r compare_naive}
q1_naive <- suppressMessages(
  slp_gas("HEF", dates_q1, temps_q1, annual_consumption = 15000)
)

cat("Reference-station KW \u2014 Q1 total:", round(sum(q1_hef$kwh)),  "kWh\n")
cat("Naive KW            \u2014 Q1 total:", round(sum(q1_naive$kwh)), "kWh\n")
```

The naive approach produces a Q1 total of exactly `annual_consumption` (15,000
kWh) because the denominator contains only winter temperatures, making $KW$
far too small. The reference-station approach correctly yields a higher Q1
total — winter months account for a disproportionate share of annual gas
consumption.

## Using your own temperature data

If the built-in stations do not adequately represent your customer's location,
you can pass custom `dates` and `temperatures` vectors. The
most straightforward source is the open-data portal of the Deutscher
Wetterdienst (DWD). The
[`rdwd`](https://cran.r-project.org/package=rdwd) package downloads and parses
DWD data with no registration or API key required:

```{r dwd_download, eval=FALSE}
library(rdwd)

# Find the download URL for Hamburg (daily climate summary, historical)
link <- selectDWD("Hamburg-Fuhlsbuettel", res = "daily", var = "kl",
                  per = "historical")
file <- dataDWD(link, read = FALSE)
raw  <- readDWD(file, varnames = FALSE)

# Keep one full calendar year; TMK is the daily mean temperature in degrees C
raw_2024 <- raw[format(raw$MESS_DATUM, "%Y") == "2024", ]
dates_ref <- as.Date(raw_2024$MESS_DATUM)
temps_ref <- raw_2024$TMK

kw <- slp_gas_kundenwert("HEF", dates = dates_ref, temperatures = temps_ref,
                      annual_consumption = 15000)
```

The `TMK` column is the arithmetic daily mean temperature — this is the
*simple daily mean* variant of the *Allokationstemperatur* as defined in
the BDEW Leitfaden. If you prefer the geometrically-weighted 4-day mean
recommended for distribution network operators, apply it to the temperature
vector before proceeding (see the main gas vignette for the formula).
