---
title: "Generate a standard load profile for gas"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generate a standard load profile for gas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  dpi       = 300,
  fig.asp   = 0.618,
  fig.width = 6,
  fig.height = 3,
  fig.align = "center",
  out.width = "90%"
)
```

```{=html}
<style>
  .content h3 {
    margin-top: -30px !important;
  }

  details {
    margin-bottom: 40px;
  }

  table {
    width: 100%;
    table-layout: auto;
  }

  table col {
    width: auto !important;
  }
</style>
```

``` {r, message=FALSE, include=FALSE}
library(standardlastprofile)
library(ggplot2)
```

A standard load profile reflects the typical pattern of gas consumption. It
is used for forecasting and balancing purposes for various customer types or
customer groups for which no continuous measurement data is available. In this
respect, standard load profiles represent a simplification. 

In this vignette we explain how `slp_gas()` implements the algorithm of the
**synthetic procedure** (German: *synthetisches Verfahren*), following the
definitions by BDEW, VKU, and GEODE in
the *Leitfaden Abwicklung von Standardlastprofilen Gas* (2025).[^leitfaden]

[^leitfaden]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin. Available at <https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>

## Synthetic procedure

The synthetic procedure takes a bottom-up approach: the gas consumption for a
gas network is explained by the sum of the individual gas consumers within that
network.^[In addition to the synthetic procedure, there is also the analytical
procedure. In brief, under this procedure, a network operator forecasts
the proportion of the daily gas volume attributable to *all* SLP customers
based on the expected temperature for that day and the residual gas amount
derived from the total gas consumption, minus the gas volume attributable to
customers with metered consumption. The analytical procedure is not in scope of
this package.]

For each customer type, i.e. profile, the gas
consumption of day $D$ is the result of three factors:

1. outside temperature
2. preferences and habits
3. the day of the week

Mathematically, this is expressed as:[^procedure]

$$
Q(D) = h(\vartheta_D) \times KW \times F_{WT, D}
$$
where:

* $h(\ldots)$ is a profile function evaluated at the daily
  temperature $\vartheta_D$.
* $KW$ (German: "Kundenwert") — a customer-specific scaling factor in kWh/day.
* $F_{WT}$ is a weekday factor that adjusts for consumption differences across
  days of the week.

[^procedure]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Allgemeine Begriffsbestimmungen / Erläuterungen, p. VIII

We will discuss these three factors in more detail below.

### Profile function and temperature

Gas consumption is strongly correlated with outdoor temperature; the colder
it gets, the more gas is consumed. The relationship between outside
temperature and gas consumption is captured by the profile function $h(\vartheta)$. 

This function was initially published in 2003 as an S-shaped sigmoid function:[^Hellwig2003] 

$$
h(\vartheta) = \frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D
$$

Since then, it has been revised as part of an ongoing debate. In a
research paper from 2015 it was found that a pure sigmoid approach
systematically underestimated consumption at very cold temperatures. The authors
proposed an improved formulation called **SigLinDe**
(**Sig**moid + **Lin**ear + **De**utschland).[^ffe2015] SigLinDe adds
a linear component to that sigmoid equation representing space-heating demand
(German: *Heizgas-Gerade*) and hot water demand (German: *Warmwasser-Gerade*).

$$
h(\vartheta) = 
\underbrace{\frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D}_{\text{sigmoid part}}
+
\underbrace{\max\!\left(m_H \vartheta + b_H,\; m_W \vartheta + b_W\right)}_{\text{linear part}}
$$

```{r siglinde-toggle-chart, echo=FALSE, results='asis', eval=knitr::is_html_output()}
cat('
<div id="siglinde-widget" style="max-width: 600px; margin: 1.5em auto; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div style="margin-bottom: 12px;">
<div style="font-size: 15px; font-weight: 600; color: #1a1a1a;">Higher h(\u03d1) means more gas consumption on that day</div>
<style>
.sig-toggle-label {
  display: inline-block;
  position: relative;
}
.sig-toggle-label > span {
  position: relative;
}
.sig-toggle-label::before {
  content: attr(data-text);
  font-weight: 700;
  visibility: hidden;
  display: block;
  height: 0;
  overflow: hidden;
  pointer-events: none;
}
</style>
<p style="font-size: 13px; color: #555; margin: 4px 0 0; line-height: 1.4;">Toggle between <span id="siglinde-label-sig" class="sig-toggle-label" data-text="Sigmoid"><span id="siglinde-label-sig-inner" style="font-weight: 700; color: #0CC792;">Sigmoid</span></span>
<span style="display: inline-block; vertical-align: middle; width: 2.2em; height: 1.2em; margin: 0 2px;">
<button id="siglinde-toggle" role="switch" aria-checked="false" aria-label="Toggle between a pure Sigmoid and SigLinDe"
  style="position: relative; width: 100%; height: 100%; background: #fff; outline: 2px solid #333; border: none; border-radius: 0; cursor: pointer; padding: 0;">
<span id="siglinde-thumb" style="position: absolute; top: 50%; left: 0.15em; width: 0.75em; height: 0.75em; background: #333; border-radius: 0; display: block; transition: transform 0.2s ease; transform: translateY(-50%);"></span>
</button>
</span>
and <span id="siglinde-label-sld" class="sig-toggle-label" data-text="SigLinDe"><span id="siglinde-label-sld-inner" style="font-weight: 400;">SigLinDe</span></span> functions</p>
</div>
<div id="siglinde-chart"></div>
</div>

<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const A = 1.0443538, B = -35.0333754, C = 6.2240634, D = 0.0502917, theta0 = 40;
const mH = -0.053583, bH = 0.9995901, mW = -0.0021758, bW = 0.1633299;

function sigmoid(t) {
  return A / (1 + Math.pow(B / (t - theta0), C)) + D;
}
function linear(t) {
  return Math.max(mH * t + bH, mW * t + bW);
}

const temps = [];
for (let t = -15; t <= 25; t += 0.25) temps.push(t);

const sigOnly  = temps.map(t => sigmoid(t));
const siglinde = temps.map(t => sigmoid(t) + linear(t));

const chartEl  = document.getElementById("siglinde-chart");
const toggle   = document.getElementById("siglinde-toggle");
const thumb    = document.getElementById("siglinde-thumb");
const labelSig = document.getElementById("siglinde-label-sig-inner");
const labelSld = document.getElementById("siglinde-label-sld-inner");

let active = false;
let blend  = 0;
let animId = null;

function render(t) {
  const data = temps.map((temp, i) => ({
    temp,
    h: sigOnly[i] + t * (siglinde[i] - sigOnly[i])
  }));

  const marks = [
    Plot.lineY(data, {
      x: "temp", y: "h",
      stroke: "#0CC792", strokeWidth: 2.5
    })
  ];

  const plot = Plot.plot({
    width: 600,
    height: 340,
    marginLeft: 50,
    marginBottom: 40,
    x: {
      label: "Temperature (\u00b0C)",
      tickFormat: d => d + "\u00b0",
      grid: true
    },
    y: {
      label: "h(\u03d1)",
      domain: [0, 3],
      grid: true
    },
    marks,
    style: {
      fontSize: "13px",
      fontFamily: "-apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif",
      background: "transparent"
    }
  });

  chartEl.innerHTML = "";
  chartEl.appendChild(plot);
}

function animate() {
  const target = active ? 1 : 0;
  const step = 0.06;
  const diff = target - blend;
  if (Math.abs(diff) < step) {
    blend = target;
    render(blend);
    animId = null;
    return;
  }
  blend += Math.sign(diff) * step;
  render(blend);
  animId = requestAnimationFrame(animate);
}

toggle.addEventListener("click", () => {
  active = !active;
  toggle.setAttribute("aria-checked", active);
  thumb.style.transform = active ? "translateY(-50%) translateX(1.1em)" : "translateY(-50%)";
  labelSig.style.fontWeight = active ? "400" : "700";
  labelSig.style.color = active ? "" : "#0CC792";
  labelSld.style.fontWeight = active ? "700" : "400";
  labelSld.style.color = active ? "#0CC792" : "";
  if (animId) cancelAnimationFrame(animId);
  animId = requestAnimationFrame(animate);
});

render(0);
</script>
')
```

The SigLinDe approach was adopted by BDEW in 2018 and has since been the binding standard.[^bdew2018]
`slp_gas()` implements the SigLinDe approach. The parameters $A$,
$B$, $C$, $D$, $\vartheta_0$, $m_H$, $b_H$, $m_W$, $b_W$ are profile-specific
constants.[^params_source]

[^params_source]: The coefficients A, B, C, D, m~H~, b~H~, m~W~, b~W~ for each consumption profile are published in Anlage 6 of the BDEW Leitfaden (pp. 140–163).

[^Hellwig2003]: Hellwig, Mark (2003). *Entwicklung und Anwendung parametrisierter Standard-Lastprofile*. Dissertation, TU München.

[^ffe2015]: Hinterstocker, M., Eberl, B., von Roon, S. (2015). *Weiterentwicklung des Standardlastprofilverfahrens Gas*. Endbericht. Forschungsgesellschaft für Energiewirtschaft mbH (FfE), available at <https://www.bdew.de/media/documents/201507_Weiterentwicklung-SLP-Gas.pdf>

[^bdew2018]: BDEW/VKU/GEODE-Leitfaden Abwicklung von Standardlastprofilen Gas, Stand 28.10.2025, available at <https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>

The result of $h(\vartheta)$ is a number > 0 without a dimension:

``` {r h-vartheta-minimal-example, message=FALSE}
# profile "GMF", Ausprägung: 34
slp_gas_siglinde(
  theta  =   8.0,
  A      =   1.0443538,
  B      = -35.0333754,
  C      =   6.2240634,
  D      =   0.0502917,
  theta0 =  40.0,
  mH     =  -0.053583,
  bH     =   0.9995901,
  mW     =  -0.0021758,
  bW     =   0.1633299
)
```

The output of this function will be scaled to kWh once we add $KW$ into the equation.  

### Customer value 'KW'

The customer value takes into account a customer’s individual consumption behaviour. It is a scaling factor expressed in kWh/day and expresses the average gas consumption at 8°C. We can rearrange the
gas consumption equation from above and express $KW$ mathematically as:

$$
KW = \frac{E_a}{\displaystyle\sum_{D} h(\vartheta_D) \times F_{WT,D}}
$$

Where 

* $E_a$ is the gas consumption (for a certain period of time, usually one year)
* $\vartheta_D$ is the profile function value for day $D$
* $F_{WT,D}$ is the scaling factor for weekday $D$

Let's make this concrete on the example of a single-family house (profile ID: `HEF`)
in Hamburg. We assume that the gas consumption in the previous calendar year
(or last billing period for that matter) was 15,000 kWh. With that information
we can calculate $KW$. We know $\vartheta_D$ from the previous section and we can
fetch the temperature data for Hamburg from the [German Meteorological Service (DWD)](https://www.dwd.de/EN/Home/home_node.html).
If we put all this information together into the equation above, this is the result:

```{r kundenwert_table, echo=FALSE, message=FALSE}
# Hamburg built-in reference temperatures (WMO climate normal 1991-2020)
# Uses the full 366-day climatological series, matching slp_gas_kundenwert()
ref <- slp_gas_normtemperatur
ref <- ref[ref$station == "Hamburg", c("date", "temp_c_mean")]

# Compute h(theta) for HEF, variant 34
p <- standardlastprofile:::.gas_profile_params[["34"]][["HEF"]]
ref$h <- slp_gas_siglinde(
  ref$temp_c_mean, p$A, p$B, p$C, p$D, p$theta0,
  p$mH, p$bH, p$mW, p$bW
)

# Weekday factors (all 1 for HEF, but include for completeness)
wt_keys <- standardlastprofile:::get_gas_weekday_key(ref$date)
fwt <- standardlastprofile:::.gas_weekday_factors[["HEF"]]
ref$f_wt <- unname(fwt[wt_keys])

# Show first and last few rows with ellipsis
show <- rbind(
  head(ref, 1),
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  ref[floor(nrow(ref) / 2) + c(-1, 0, 1), ],
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  tail(ref, 1)
)
show_df <- data.frame(
  Date = ifelse(is.na(show$date), "\u22ee", format(show$date, "%b %e")),
  `°C` = ifelse(is.na(show$temp_c_mean), "\u22ee", round(show$temp_c_mean, 1)),
  `h(ϑ)` = ifelse(is.na(show$h), "\u22ee", round(show$h, 4)),
  check.names = FALSE
)
knitr::kable(show_df, align = "lrr")
```

The sum in the denominator across all `r nrow(ref)` days gives
$\sum_D h(\vartheta_D) \times F_{WT,D}$ = `r round(sum(ref$h * ref$f_wt), 2)`,
so $KW$ = 15,000 / `r round(sum(ref$h * ref$f_wt), 2)` =
**`r round(15000 / sum(ref$h * ref$f_wt), 2)` kWh/day**.

We have covered two of the three factors of the equation to predict gas
consumption for any day $D$ – $Q(D) = KW \times h(\vartheta_D) \times F_{WT, D}$ –;
the one missing is the weekday factor $F_{WT, D}$.

### Weekday factors

Unlike the electricity profiles, which group Monday–Friday into a single
`workday` type, the gas procedure uses a separate scaling factor for every weekday 
from Monday through Sunday.[^fwt]

Public holidays are treated as a Sunday — the Leitfaden notes that
consumption on public holidays most closely resembles a Sunday.[^holiday_rule]
24 December and 31 December are treated as Saturday unless they fall
on a Sunday; this convention mirrors the established rule from the BDEW
electricity SLP procedure and is applied here by analogy.

[^holiday_rule]: BDEW/VKU/GEODE (2018). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Anlage 3, pp. 129–130.

[^fwt]: BDEW/VKU/GEODE (2018). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Anlage 6, pp. 140–163 (per-profile data sheets).

For the residential profiles `HEF`, `HMF`, and `HKO`, however, all weekday factors
are equal to 1, that is, gas consumption in households is assumed not to differ
significantly by day of the week. 
Commercial profiles on the other hand show visible weekday patterns, the range of
the weekday factors is `r do.call(rbind, standardlastprofile:::.gas_weekday_factors) |> range() |> paste(collapse = " - ")`,
illustrated by the chart below:

```{r weekday_factors_data, echo=FALSE}
days <- c("Mo", "Tu", "We", "Th", "Fr", "Sa", "Su")
profiles_f <- names(standardlastprofile:::.gas_weekday_factors)

fwt <- do.call(rbind, lapply(profiles_f, \(pid) {
  wfct <- standardlastprofile:::.gas_weekday_factors[[pid]]
  data.frame(
    profile_id = pid,
    day        = factor(names(wfct), levels = days),
    fct_value  = unname(wfct),
    category   = if(startsWith(pid, "H")) "household" else "commercial"
  )
}))
```

```{r weekday_factors_heatmap_static, echo=FALSE, fig.asp=1, fig.retina=2, eval=!knitr::is_html_output()}
ggplot(fwt, aes(day, profile_id, fill = fct_value - 1)) +
  geom_tile(width = .95, height = .95) +
  coord_equal(ratio = .75) +
  scale_fill_gradient2(NULL, labels = scales::label_percent(), midpoint = 0, low = "#311dfc", mid = "#fffafe", high = "#ff005e") +
  facet_grid(rows = vars(category), scales = "free", space = "free") +
  labs(
    title   = NULL,
    caption = NULL,
    x       = NULL,
    y       = NULL,
    fill    = NULL
  ) +
  theme_minimal() +
  theme(legend.position    = "bottom") +
  theme(legend.key.width = unit(0.099,"npc")) +
  theme(panel.grid = element_blank())
```

```{r weekday_factors_heatmap_html, echo=FALSE, results='asis', eval=knitr::is_html_output()}
fwt_json <- jsonlite::toJSON(
  data.frame(
    profile_id = fwt$profile_id,
    day        = as.character(fwt$day),
    fct_value  = fwt$fct_value,
    category   = fwt$category
  ),
  dataframe = "rows"
)
# Profile descriptions for y-axis tooltips
desc_pids <- unique(as.character(fwt$profile_id))
desc_map <- vapply(desc_pids, \(pid) {
  slp_info(pid)[[pid]]$description
}, character(1))
desc_json <- jsonlite::toJSON(as.list(desc_map), auto_unbox = TRUE)
cat(sprintf('
<div id="fwt-widget" style="max-width: 600px; margin: 1.5em auto; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div id="fwt-chart"></div>
</div>
<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const data = %s;
const descriptions = %s;
const days = ["Mo","Tu","We","Th","Fr","Sa","Su"];
const householdIds = [...new Set(data.filter(d => d.category === "household").map(d => d.profile_id))];
const commerceIds  = [...new Set(data.filter(d => d.category === "commercial").map(d => d.profile_id))];
const profileOrder = [...householdIds, ...commerceIds];

const chart = Plot.plot({
  marginLeft: 40,
  marginBottom: 30,
  height: profileOrder.length * 32 + 60,
  x: { domain: days, padding: 0.08, label: null },
  y: { domain: profileOrder, padding: 0.08, label: null },
  color: {
    type: "diverging",
    domain: [-0.6, 0.3],
    pivot: 0,
    range: ["#311dfc", "#fffafe", "#ff005e"],
    label: "Weekday factors for SLP gas profiles as deviation from 1",
    tickFormat: "+.0%%",
    legend: true
  },
  marks: [
    Plot.cell(data, {
      x: "day",
      y: "profile_id",
      fill: d => d.fct_value - 1,
      tip: true,
      title: d => {
        const v = d.fct_value - 1;
        return v === 0 ? "0%%" : (v > 0 ? "+" : "") + (v * 100).toFixed(1) + "%%";
      }
    }),
    Plot.ruleY([householdIds.length - 0.5], { stroke: "#999", strokeDasharray: "3,3" })
  ]
});

document.getElementById("fwt-chart").appendChild(chart);

// Add tooltips to y-axis tick labels
chart.querySelectorAll("text").forEach(el => {
  const pid = el.textContent.trim();
  if (!descriptions[pid]) return;
  el.style.cursor = "help";
  const t = document.createElementNS("http://www.w3.org/2000/svg", "title");
  t.textContent = descriptions[pid];
  el.prepend(t);
});
</script>
', fwt_json, desc_json))
```

## Usage

With the algorithm covered, let us see how to use `slp_gas()` in practice.

### A minimal working example

The minimal input is a `profile_id`, a `dates` vector, and a `temperatures`
vector. By default `kundenwert` is `NULL` and `annual_consumption` is `1000`.
Below we use a short synthetic temperature series to illustrate the output
format:

```{r basic_usage, message=FALSE}
dates <- seq.Date(as.Date("2026-01-12"), as.Date("2026-01-18"), by = "day")
temps <- c(-3.2, -1.8, 0.4, 2.1, 4.5, 3.8, 1.2)

result <- slp_gas(
  profile_id = "HEF",
  dates = dates,
  temperatures = temps,
  kundenwert = NULL,        # default
  annual_consumption = 1000 # default
  )
result
```

> When `kundenwert = NULL` (the default), `slp_gas()` derives $KW$ automatically
> from `annual_consumption` and the supplied temperature series. 

Without `kundenwert` and without temperature data that covers a whole year
at least chances are that the result is off – by a lot. The minimal example above
shall only illustrate the output and below we will discuss a more realistic
example.

### A full-year example

To produce a meaningful result the temperature series should cover a full
year at least. In practice you would use measured data from a trustworthy
institution like the German Meteorological Service, or DWD for short. The
[`rdwd`](https://cran.r-project.org/package=rdwd) package provides a
straightforward interface to the open-data portal of the DWD 
(Deutscher Wetterdienst).[^dwd] No registration or API key is required.

[^dwd]: DWD Open Data: <https://opendata.dwd.de/>

#### rdwd package 

The workflow has three steps:

1. `selectDWD()` — find the download URL for a station and dataset.
2. `dataDWD()` — download and unzip the file.
3. `readDWD()` — parse the file into a data frame.

For gas SLPs, we need the **daily climate summary** (`var = "kl"`,
`res = "daily"`). The column `TMK` contains the daily mean temperature in °C.

### Comparing three climate zones

Germany spans several distinct climate zones that lead to meaningfully
different gas consumption patterns:

* **Freiburg im Breisgau** — mild oceanic-continental climate in the Upper
  Rhine Plain, sheltered by the Black Forest; the warmest and sunniest large
  city in Germany, with comparatively low heating demand.
* **Hamburg** — maritime climate on the North Sea coast; mild winters but
  persistent cloud cover and wind, leading to moderate heating demand.
* **Chemnitz** — continental climate in the Erzgebirge foothills of Saxony;
  colder winters and greater temperature amplitude than the other two cities,
  resulting in the highest heating demand of the three.

The example below computes gas consumption for three representative profiles —
`HEF` (single-family home), `GBD` (services), and `GWA` (laundries). We assume
a yearly gas consumption of 15.000 kWh for all profiles and cities. With
`rdwd` installed, replace the embedded temperature series with real DWD station
data (see code comment):

```{r dwd_cities_download, eval=FALSE}
# Download real 2024 data with rdwd (requires internet access)
library(rdwd)

get_temps_dwd <- \(station_name) {
  link <- selectDWD(station_name, res = "daily", var = "kl", per = "historical")
  file <- dataDWD(link, read = FALSE)
  raw  <- readDWD(file, varnames = FALSE)
  raw  <- raw[format(raw$MESS_DATUM, "%Y") == "2024", ]
  list(
    dates = as.Date(raw$MESS_DATUM),
    temps = raw$TMK
  )
}

temps_list <- list(
  "Freiburg im Breisgau" = get_temps_dwd("Freiburg"),
  "Hamburg"              = get_temps_dwd("Hamburg-Fuhlsbüttel"),
  "Chemnitz"             = get_temps_dwd("Chemnitz")
)
```

```{r dwd_cities, fig.asp=0.9, fig.retina=2, message=FALSE}
#| fig.alt = "Facet grid showing monthly gas consumption across three German
#|  climate zones (Freiburg, Hamburg, Chemnitz) for three gas profiles
#|  (HEF, GBD, GWA). Chemnitz shows the highest winter peaks."

# Climate-representative temperature series (2024, daily mean °C).
# Parameters are calibrated to long-term DWD station normals:
#   mean  = annual mean temperature
#   amp   = seasonal amplitude (peak-to-mean)
#   phase = day-of-year offset of the coldest point (~Jan 15)
set.seed(2024)
dates_2024 <- seq.Date(as.Date("2024-01-01"), as.Date("2024-12-31"), by = "day")
doy        <- as.integer(format(dates_2024, "%j"))

make_temps <- \(mean, amp, phase = 15, sd = 2.5) {
  mean - amp * cos(2 * pi * (doy - phase) / 365) + rnorm(366, sd = sd)
}

# Freiburg: warm (~11.5 °C mean), moderate amplitude — sheltered Upper Rhine
# Hamburg:  cool-temperate (~10 °C mean), damped amplitude — maritime North Sea
# Chemnitz: coldest (~9 °C mean), largest amplitude — continental Saxony
temps_list <- list(
  "Freiburg im Breisgau" = list(dates = dates_2024, temps = make_temps(11.5,  9.5)),
  "Hamburg"              = list(dates = dates_2024, temps = make_temps(10.0, 10.0)),
  "Chemnitz"             = list(dates = dates_2024, temps = make_temps( 9.0, 11.5))
)

profiles <- c("HEF", "GBD", "GWA")

gas_cities <- do.call(rbind, Map(\(city_data, city) {
  out      <- slp_gas(profiles, city_data$dates, city_data$temps, annual_consumption = 15000)
  out$city <- city
  out
}, temps_list, names(temps_list)))

# Aggregate to monthly totals for a cleaner visual
gas_cities$month <- as.Date(format(gas_cities$date, "%Y-%m-01"))

gas_monthly <- aggregate(
  kwh ~ city + profile_id + month,
  data = gas_cities,
  FUN  = sum
)

# Order cities warm-to-cold (left-to-right in columns)
gas_monthly$city <- factor(
  gas_monthly$city,
  levels = c("Freiburg im Breisgau", "Hamburg", "Chemnitz")
)

library(ggplot2)

ggplot(gas_monthly, aes(month, kwh, fill = city)) +
  geom_col(show.legend = FALSE) +
  facet_grid(city ~ profile_id) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b") +
  scale_y_continuous("Monthly gas consumption (kWh)") +
  scale_fill_manual(values = c(
    "Freiburg im Breisgau" = "#FA9529",
    "Hamburg"              = "#0CC792",
    "Chemnitz"             = "#961BFA"
  )) +
  labs(
    title    = "Gas standard load profiles across three German climate zones",
    subtitle = "Annual consumption normalised to 1,000 kWh/a \u2022 illustrative temperature series",
    caption  = "Profile parameters: BDEW Leitfaden 2018",
    x        = NULL
  ) +
  theme_minimal() +
  theme(
    strip.text         = element_text(face = "bold"),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid = element_line(
      linetype = "12", lineend = "round", colour = "#FAF6F4"
    )
  )
```

The facet grid reveals two orthogonal effects:

* **Rows (profile)**: `HEF` shows the strongest winter peak because all gas
  consumption is heating-driven. `GBD` (offices and services) has a smaller
  but still marked winter peak. `GWA` (laundries) is the flattest, as
  industrial washing demand varies little with temperature and its weekday
  factors strongly suppress weekend consumption.
* **Columns (city)**: Chemnitz consistently shows higher winter consumption
  than Hamburg, which in turn exceeds Freiburg — a direct reflection of the
  colder continental climate in Saxony versus the mild Upper Rhine Plain.

### Computing the geometrically-weighted Allokationstemperatur

If you have raw daily mean temperatures from DWD and want to use the
geometrically-weighted 4-day mean recommended by BDEW, apply it before calling
`slp_gas()`. (If you already have GPT values from a meteorological
service provider, skip this step — GPT is not this formula.)

```{r allokationstemperatur, eval=FALSE}
# Geometrically weighted 4-day mean:
# (T_D + 0.5*T_{D-1} + 0.25*T_{D-2} + 0.125*T_{D-3}) / 1.875
weight_allok <- c(1, 0.5, 0.25, 0.125)

apply_allok <- \(city_data) {
  theta <- stats::filter(
    city_data$temps,
    filter = weight_allok / sum(weight_allok),
    method = "convolution",
    sides  = 1                        # only past values
  )
  keep <- !is.na(theta)
  list(dates = city_data$dates[keep], temps = as.numeric(theta[keep]))
}

temps_list_weighted <- lapply(temps_list, apply_allok)
```

### Discovering stations

To find the exact station name to pass to `selectDWD()`, use `findID()` or
open the interactive station map:

```{r dwd_stations, eval=FALSE}
rdwd::findID("Freiburg")             # returns matching station names and IDs
rdwd::selectDWD(mapDWD = TRUE)       # opens an interactive map in the browser
```

## Profile IDs

There are 15 gas profile IDs defined in the BDEW Leitfaden 2018, split into
residential and commercial / industrial types.

### Residential profiles

Residential profiles share two important characteristics: all weekday factors
equal 1 (no day-of-week differentiation), and heating load is the dominant
driver of demand variability.

| Profile | Description | Notes |
|---------|-------------|-------|
| `HEF`   | Single-family home (Einfamilienhaus) | Strongest seasonal swing; includes space heating + hot water |
| `HMF`   | Multi-family home (Mehrfamilienhaus) | Lower per-unit demand than `HEF` due to shared building envelope |
| `HKO`   | Cooking and hot water only (Kochen / Warmwasser) | No space heating; TUM profile — sigmoid only, linear part is zero |

`HKO` is appropriate for customers whose gas connection serves only a cooker
or instantaneous water heater. Its flat, temperature-insensitive shape is
visible in the SigLinDe curve plot above.

### Commercial and industrial profiles

Commercial profiles have profile-specific weekday factors that reflect the
operating hours and schedules of each sector. They are grouped here by
consumption pattern:

**Process-heat dominated** (relatively flat seasonal pattern, strong weekday
structure):

| Profile | Description |
|---------|-------------|
| `GWA`   | Laundries — large hot-water demand, insensitive to outdoor temperature |
| `GBA`   | Bakeries — continuous baking process, pronounced weekday factors |
| `GPD`   | Paper and printing — industrial process heat |

**Space-heating dominated** (strong winter peak, moderate weekday variation):

| Profile | Description |
|---------|-------------|
| `GKO`   | Small commercial (Kleinstgewerbe) |
| `GHA`   | Trade and commerce (Handel) |
| `GBD`   | Services (Dienstleistung) |
| `GBH`   | Accommodation (Beherbergung) |
| `GGA`   | Gastronomy (Gastronomie) |
| `GMK`   | Metal and automotive (Metall / Kfz) |
| `GGB`   | Mixed commercial (gemischtes Gewerbe) |
| `GMF`   | Large multi-family / mixed use (Mehrfamilienhaus groß) |

**Aggregate / summary profile**:

| Profile | Description |
|---------|-------------|
| `GHD`   | Trade, commerce and services aggregate (GHD-Stützpunkt) — weighted mean across all G-types |

`GHD` is used when a customer cannot be assigned to a specific sector. Its
parameters are a weighted average of the individual commercial profiles and
its weekday factors are close to but not equal to 1.



### Customer value (Kundenwert)

The customer value $KW$ scales the dimensionless values $h(\vartheta)$ to
match a specific customer's consumption.[^kw] It indicates the daily
consumption of a standard load profile customer at a temperature of 8°C
This is where the $h(\vartheta)$ returns 1 for all profiles. This is how
the SigLinDe profiles been specified.

[^kw]: BDEW/VKU/GEODE (2018). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Abschnitt 3.6.1, p. 29; full derivation with worked examples in Anlage 1, pp. 116–122.

When `kundenwert = NULL` (the default), `slp_gas()` derives $KW$
automatically from `annual_consumption` and the supplied temperature series.
You can also supply a fixed `kundenwert` directly, which is useful when you
already know the customer's daily baseline and want to bypass the annual
normalisation.


### 

*Allokationstemperatur* $\vartheta_D$ summarises the temperature for a given
day. Two approaches are
defined in the Leitfaden:[^allok]

**Simple daily mean** — the Leitfaden specifies an arithmetic mean of 24 hourly
values covering the gas day, without writing out an explicit summation formula.
In standard notation this is:

$$\vartheta_D = \frac{1}{24} \sum_{i=1}^{24} T_i$$

**Geometrically-weighted 4-day mean** — recommended by BDEW for grid
operators, giving more weight to the current day. The document writes the
denominator in expanded form:

$$\vartheta_D = \frac{T_D + 0.5 \cdot T_{D-1} + 0.25 \cdot T_{D-2} + 0.125 \cdot T_{D-3}}{1 + 0.5 + 0.25 + 0.125}$$

[^allok]: BDEW/VKU/GEODE (2018). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Abschnitt 3.5.2, p. 22.

`slp_gas()` accepts whichever values you pass in `temperatures` — the
choice of method is the user's responsibility
(see [Real temperature data from DWD](#real-temperature-data-from-dwd)).

## The SigLinDe profile function

The profile function $h(\vartheta)$ is the mathematical heart of the synthetic
procedure. It maps a temperature to a dimensionless load factor. The SigLinDe
function comes in two variants (German: *Ausprägungen*) — **34** (57 % linear
component, steeper heating slope) and **33** (45 % linear component) — both
introduced in the 2015 FfE report.[^siglinde] The BDEW Leitfaden 2018
recommends that distribution network operators test both and select the better fit for their
grid. `slp_gas()` uses variant 34 by default and lets you switch
with `variant = "33"`.

$$
h(\vartheta) =
\underbrace{\frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D}_{\text{sigmoid part}}
+
\underbrace{\max\!\left(m_H \vartheta + b_H,\; m_W \vartheta + b_W\right)}_{\text{linear part}}
$$

The **sigmoid part** captures the temperature-dependent space-heating demand
that dominates in winter but flattens in summer. The **linear part** adds two
straight lines — a space-heating line (German: *Heizgas-Gerade*, $m_H$, $b_H$)
and a hot water line (German: *Warmwasser-Gerade*, $m_W$, $b_W$) — representing
supplementary heating and domestic hot water loads. 

[^siglinde]: BDEW/VKU/GEODE (2018). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Anlage 7, pp. 164–166.

We can inspect the profile function for a selection of
profiles across the temperature range that is relevant in Germany:

```{r siglinde_curve, echo=TRUE, fig.asp=0.618, fig.retina=2}
library(standardlastprofile)
library(ggplot2)

theta <- seq(-15, 25, by = 0.5)

profiles <- c("HEF", "HMF", "GKO", "GHA")
curves <- do.call(rbind, lapply(profiles, \(pid) {
  p <- standardlastprofile:::.gas_profile_params[["34"]][[pid]]
  h <- slp_gas_siglinde(
    theta, p$A, p$B, p$C, p$D, p$theta0,
    p$mH, p$bH, p$mW, p$bW
  )
  data.frame(profile_id = pid, theta = theta, h = h)
}))

ggplot(curves, aes(theta, h, colour = profile_id)) +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey70", linewidth = 0.4) +
  geom_vline(xintercept = 8, linetype = "dotted", colour = "grey70", linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(labels = \(x) paste0(x, "°C"), breaks = c(-10, 0, 8, 20)) +
  scale_y_continuous("h(ϑ)") +
  labs(
    title    = "SigLinDe profile functions",
    subtitle = "Higher h(ϑ) means more gas consumption on that day",
    caption  = "data: BDEW Leitfaden 2018, Anlage 6",
    colour   = NULL,
    x        = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position    = "top",
    panel.grid = element_line(
      linetype = "12", lineend = "round", colour = "#FAF6F4"
    )
  )
```

At low temperatures all profiles show a high load factor because of space
heating demand. As temperature rises, $h(\vartheta)$ decreases and flattens.
`HKO` has no space-heating component and therefore shows a much lower and
flatter curve.
