---
title: "SigLinDe Parameters and Weekday Factors for Gas SLP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gas SLP — SigLinDe Parameters and Weekday Factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(standardlastprofile)
```

This vignette provides machine-readable reference tables for all parameters used
by `slp_gas()` to compute daily gas consumption under the BDEW/VKU/GEODE
standard load profile procedure.

## Source

All values are taken from the

> **BDEW/VKU/GEODE Leitfaden — Abwicklung von Standardlastprofilen Gas**,
> Kooperationsvereinbarung Gas, Anlage XIV.2,
> Stand vom 28.10.2025, Anhang 6, pages 145–166.

The document is available at
<https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>.

## Profile overview

The BDEW Leitfaden defines **15 gas standard load profiles**. Three are
residential (prefix 'H' for "Haushalt") and twelve are commercial or industrial
 (prefix 'G' for "Gewerbe").

```{r profile_overview}
pids <- c("HEF", "HMF", "HKO",
           "GKO", "GHA", "GMK", "GBD", "GBH", "GWA",
           "GGA", "GBA", "GGB", "GPD", "GMF", "GHD")
info <- slp_info(pids)
info_de <- slp_info(pids, language = "DE")
overview <- data.frame(
  Profile     = pids,
  Description = vapply(pids, \(p) info[[p]]$description, character(1)),
  German      = vapply(pids, \(p) info_de[[p]]$description, character(1)),
  stringsAsFactors = FALSE, row.names = NULL
)
knitr::kable(overview, caption = "Table 1: BDEW gas standard load profile identifiers and descriptions.")
```

## SigLinDe coefficients

Daily gas consumption is computed using the **SigLinDe** method
(Sigmoid + Linear + Deutschland). The dimensionless daily heating demand
$h(\vartheta)$ for an allocation temperature $\vartheta$ (in °C) is:

$$h(\vartheta) = \frac{A}{1 + \left(\frac{B}{\vartheta - \vartheta_0}\right)^C} + D + \max\left(m_H \cdot \vartheta + b_H,\; m_W \cdot \vartheta + b_W\right)$$

Each profile has its own set of nine coefficients. The linear component
$\max(m_H \vartheta + b_H,\; m_W \vartheta + b_W)$ represents the envelope of
a space-heating line (*Heizgas-Gerade*) and a hot-water line
(*Warmwasser-Gerade*). The parameters differ between
**variant 34** (57 % linear component, default) and **variant 33**
(45 % linear component).

### Variant 34 (Ausprägung 34)

```{r params_34}
params_34 <- standardlastprofile:::.gas_profile_params[["34"]]
df_34 <- do.call(rbind, lapply(names(params_34), \(pid) {
  p <- params_34[[pid]]
  data.frame(
    Profile = pid,
    A       = p$A,
    B       = p$B,
    C       = p$C,
    D       = p$D,
    theta_0 = p$theta0,
    m_H     = p$mH,
    b_H     = p$bH,
    m_W     = p$mW,
    b_W     = p$bW,
    stringsAsFactors = FALSE
  )
}))
knitr::kable(
  df_34, digits = 7, row.names = FALSE,
  caption = "Table 2: SigLinDe coefficients — variant 34 (Ausprägung 34, 57 % linear component). Source: BDEW Leitfaden, Anhang 6, pp. 145–166."
)
```

### Variant 33 (Ausprägung 33)

```{r params_33}
params_33 <- standardlastprofile:::.gas_profile_params[["33"]]
df_33 <- do.call(rbind, lapply(names(params_33), \(pid) {
  p <- params_33[[pid]]
  data.frame(
    Profile = pid,
    A       = p$A,
    B       = p$B,
    C       = p$C,
    D       = p$D,
    theta_0 = p$theta0,
    m_H     = p$mH,
    b_H     = p$bH,
    m_W     = p$mW,
    b_W     = p$bW,
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}))
knitr::kable(
  df_33, digits = 7, row.names = FALSE,
  caption = "Table 3: SigLinDe coefficients — variant 33 (Ausprägung 33, 45 % linear component). Source: BDEW Leitfaden, Anhang 6, pp. 145–166."
)
```

### Notes

- `HKO` (cooking and hot water only) has $m_H = b_H = m_W = b_W = 0$: the
  linear part is zero and only the sigmoid remains.
- `HMF` and `GMF` share identical coefficients in both variants — `GMF` (large
  multi-family / mixed use) uses the residential multi-family curve.
- $\vartheta_0 = 40\,°C$ for all profiles (the pole temperature of the
  sigmoid).

## Weekday factors ($F_{WT}$)

The weekday factor $F_{WT}$ scale daily consumption for day-of-week effects.
Residential profiles (`HEF`, `HMF`, `HKO`) have all factors equal to 1 — no
weekday differentiation.

```{r weekday_factors}
fwt <- standardlastprofile:::.gas_weekday_factors
df_fwt <- do.call(rbind, lapply(names(fwt), \(pid) {
  f <- fwt[[pid]]
  data.frame(
    Profile = pid,
    Mo = f[["Mo"]],
    Tu = f[["Tu"]],
    We = f[["We"]],
    Th = f[["Th"]],
    Fr = f[["Fr"]],
    Sa = f[["Sa"]],
    Su = f[["Su"]],
    stringsAsFactors = FALSE,
    row.names = NULL
  )
}))
knitr::kable(
  df_fwt, digits = 4, row.names = FALSE,
  caption = "Table 4: Weekday factors (F_WT) for all 15 BDEW gas standard load profiles. Mo = Monday, …, Su = Sunday. Source: BDEW Leitfaden, Anhang 6, pp. 145–166."
)
```

### Notes

- Weekday factors are the same for both variant 33 and variant 34.
