---
title: "Get started"
description: "Get started with standardlastprofile - an R package for BDEW Standardlastprofile (SLP Strom and SLP Gas). Generate electricity and gas standard load profiles for the German energy market."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A **standard load profile** (German: *Standardlastprofil*, SLP) is a
representative pattern of energy consumption used by the German energy industry
to forecast demand for customer groups that are not continuously metered. The
profiles are published by the BDEW (Bundesverband der Energie- und
Wasserwirtschaft e.V.) and form the basis of balancing and settlement in the
German electricity and gas markets.

This package provides functions to generate standard load profiles for
**electricity** and **gas**.

## Electricity

`slp_electricity()` produces a 15-minute resolution load profile in watts,
normalised to an annual consumption of 1,000 kWh.

```{r electricity}
library(standardlastprofile)

G5 <- slp_electricity(
  profile_id = "G5",
  start_date = "2026-01-12",
  end_date   = "2026-01-18"
)

head(G5)
```

See `?slp_electricity` for the full list of profile IDs, and the
[electricity algorithm article](slp-electricity.html) for a detailed
explanation of the algorithm.

## Gas

`slp_gas()` produces daily gas consumption in kWh. It requires daily (mean)
temperatures and a `kundenwert` (customer value, kWh/day). The Kundenwert is
derived once from a full reference year with `slp_gas_kundenwert()`; here we
pass a known value:

```{r gas}
dates <- seq.Date(as.Date("2026-01-12"), as.Date("2026-01-18"), by = "day")
temps <- c(-3.2, -1.8, 0.4, 2.1, 4.5, 3.8, 1.2)

slp_gas(
  profile_id   = "HEF",
  dates        = dates,
  temperatures = temps,
  kundenwert   = 3.055241
)
```

See `?slp_gas` for the full list of gas profile IDs, and the
[gas algorithm article](slp-gas.html) for a detailed explanation of the
parameters and the algorithm.

## Further reading

- [Electricity SLP algorithm](slp-electricity.html) — in detail
- [Gas SLP algorithm](slp-gas.html) — SigLinDe method, Kundenwert, and
  climate zone comparison in detail
- [SigLinDe parameters](slp-gas-parameters.html) — reference tables of all
  SigLinDe coefficients and weekday factors
