---
title: "SigLinDe Parameters and Weekday Factors for Gas SLP"
description: "SigLinDe profile function coefficients (A, B, C, D, mH, bH, mW, bW) and weekday factors for all 15 BDEW Standardlastprofile Gas - variants 33 and 34, from the BDEW Leitfaden (as of 2025-10-28)."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(standardlastprofile)
```

This article provides reference tables for all parameters used by `slp_gas()`
to compute daily gas consumption under the BDEW/VKU/GEODE standard load profile
procedure. To retrieve these values programmatically, use
`slp_gas_coefficients()` for the SigLinDe coefficients and
`slp_gas_weekday_factors()` for the weekday factors.

## Source

All values are taken from the **BDEW/VKU/GEODE Leitfaden — Abwicklung von
Standardlastprofilen Gas**, pages 145–166.^[The document is available at
<https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>]

## Profile overview

The BDEW Leitfaden defines 15 gas standard load profiles. Three are
residential (prefix 'H' for "Haushalt") and twelve are commercial or industrial
 (prefix 'G' for "Gewerbe").

```{r profile_overview}
pids <- c("HEF", "HMF", "HKO",
           "GKO", "GHA", "GMK", "GBD", "GBH", "GWA",
           "GGA", "GBA", "GGB", "GPD", "GMF", "GHD")
info <- slp_info(pids)
info_de <- slp_info(pids, language = "DE")
overview <- data.frame(
  Profile     = pids,
  Description = vapply(pids, \(p) info[[p]]$description, character(1)),
  German      = vapply(pids, \(p) info_de[[p]]$description, character(1)),
  stringsAsFactors = FALSE, row.names = NULL
)
knitr::kable(overview, caption = "Table 1: BDEW gas standard load profile identifiers and descriptions.")
```

## SigLinDe coefficients

Daily gas consumption is computed using the **SigLinDe** method
(Sigmoid + Linear + Deutschland). The dimensionless daily heating demand
$h(\vartheta)$ for a daily temperature $\vartheta$ (in °C) is:

$$h(\vartheta) = \frac{A}{1 + \left(\frac{B}{\vartheta - \vartheta_0}\right)^C} + D + \max\left(m_H \cdot \vartheta + b_H,\; m_W \cdot \vartheta + b_W\right)$$

Each profile has its own set of nine coefficients. The linear component
$\max(m_H \vartheta + b_H,\; m_W \vartheta + b_W)$ represents the envelope of
a space-heating line (*Heizgas-Gerade*) and a hot-water line
(*Warmwasser-Gerade*). The parameters differ between
**variant 34** (57 % linear component, default) and **variant 33**
(45 % linear component).

Use `slp_gas_coefficients()` to retrieve these values in R:

```r
slp_gas_coefficients()                        # all 15 profiles, both variants (30 rows)
slp_gas_coefficients(variant = "34")         # all profiles, variant 34 only
slp_gas_coefficients("HEF")                  # one profile, both variants
slp_gas_coefficients("HEF", variant = "33")  # one profile, one variant
```

### Variant 34 (Ausprägung 34)

```{r params_34}
df_34 <- slp_gas_coefficients(variant = "34")
names(df_34) <- c("Profile", "Variant", "A", "B", "C", "D",
                   "theta_0", "m_H", "b_H", "m_W", "b_W")
knitr::kable(
  df_34[, -2L], digits = 7, row.names = FALSE,
  caption = "Table 2: SigLinDe coefficients — variant 34 (Ausprägung 34, 57 % linear component). Source: BDEW Leitfaden, Appendix 6, pp. 145–166."
)
```

### Variant 33 (Ausprägung 33)

```{r params_33}
df_33 <- slp_gas_coefficients(variant = "33")
names(df_33) <- c("Profile", "Variant", "A", "B", "C", "D",
                   "theta_0", "m_H", "b_H", "m_W", "b_W")
knitr::kable(
  df_33[, -2L], digits = 7, row.names = FALSE,
  caption = "Table 3: SigLinDe coefficients — variant 33 (Ausprägung 33, 45 % linear component). Source: BDEW Leitfaden, Appendix 6, pp. 145–166."
)
```

Use `slp_gas_coefficients()` to retrieve these coefficients for one or more
profiles:

```r
slp_gas_coefficients()                        # all 15 profiles, variant 34
slp_gas_coefficients("HEF", variant = "33")  # single profile, variant 33
```

### Notes

- `HKO` (cooking and hot water only) has $m_H = b_H = m_W = b_W = 0$: the
  linear part is zero and only the sigmoid remains.
- `HMF` and `GMF` share identical coefficients in both variants — `GMF` (large
  multi-family / mixed use) uses the residential multi-family curve.
- $\vartheta_0 = 40\,°C$ for all profiles (the pole temperature of the
  sigmoid).

## Weekday factors ($F_{WT}$)

The weekday factor $F_{WT}$ scales daily consumption for day-of-week effects.
Residential profiles (`HEF`, `HMF`, `HKO`) have all factors equal to 1 — no
weekday differentiation.

```{r weekday_factors}
df_long <- slp_gas_weekday_factors()
days    <- c("Mo", "Tu", "We", "Th", "Fr", "Sa", "Su")
pids    <- unique(df_long$profile_id)
df_fwt  <- do.call(rbind, lapply(pids, \(p) {
  row <- df_long[df_long$profile_id == p, ]
  as.data.frame(c(list(Profile = p), setNames(as.list(row$f_wt), days)))
}))
knitr::kable(
  df_fwt, digits = 4, row.names = FALSE,
  caption = "Table 4: Weekday factors (F_WT) for all 15 BDEW gas standard load profiles. Mo = Monday, …, Su = Sunday. Source: BDEW Leitfaden, Appendix 6, pp. 145–166."
)
```

Use `slp_gas_weekday_factors()` to retrieve these values:

```r
slp_gas_weekday_factors()        # all 15 profiles, tidy (long) format
slp_gas_weekday_factors("GWA")  # single profile
```

### Notes

- Weekday factors are the same for both variant 33 and variant 34.
