---
title: "Gas Standard Load Profiles (Standardlastprofile Gas)"
description: "BDEW Standardlastprofile Gas in R: the SigLinDe synthetic procedure, Kundenwert, and weekday factors for 15 gas profiles (HEF, HMF, HKO, GKO, and 11 more)."
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  dpi       = 300,
  fig.asp   = 0.618,
  fig.width = 6,
  fig.height = 3,
  fig.align = "center",
  out.width = "90%"
)
```

```{=html}
<style>
  .content h3 {
    margin-top: -30px !important;
  }

  details {
    margin-bottom: 40px;
  }

  table {
    width: 100%;
    table-layout: auto;
  }

  table col {
    width: auto !important;
  }
</style>
```

``` {r, message=FALSE, include=FALSE}
library(standardlastprofile)
library(ggplot2)
```

A standard load profile reflects the assumed pattern of gas consumption. It
is used for forecasting and balancing purposes for various customer types or
customer groups for which no continuous measurement data is available. In this
respect, standard load profiles represent a simplification. 

In this article we explain how `slp_gas()` implements the algorithm of the
synthetic procedure (German: "synthetisches Verfahren"), following the
definitions by BDEW, VKU, and GEODE in
the *Leitfaden Abwicklung von Standardlastprofilen Gas* (2026).[^leitfaden]

[^leitfaden]: BDEW/VKU/GEODE (2026). *Leitfaden Abwicklung von Standardlastprofilen Gas*, Kooperationsvereinbarung Gas, Annex XV, as of 2026-03-27. Berlin. Archived copy: <https://web.archive.org/web/20260619125016/https://www.bdew.de/media/documents/260327_LF_SLP_Gas_KoV_XV_CO4f7Rb.pdf> (current edition via the [BDEW gas SLP page](https://www.bdew.de/energie/standardlastprofile-gas/)).

## Synthetic procedure

The synthetic procedure takes a bottom-up approach: the gas consumption is
explained by the sum of the individual gas consumers.^[In addition to
the synthetic procedure, there is also the analytical procedure. In brief,
under this procedure, a network operator forecasts the proportion of the daily
gas volume attributable to *all* SLP customers based on the expected
temperature for that day and the residual gas amount derived from the total
gas consumption, minus the gas volume attributable to customers with metered
consumption. The analytical procedure is not in scope of this package.] The gas
consumption on day $D$ is the result of three factors:

1. outside temperature
2. preferences and habits
3. the day of the week

Mathematically, this is expressed as:[^procedure]

$$
Q(D) = h(\vartheta_D) \times KW \times F_{WT, D} \tag{1}\label{eq:qdaily}
$$
where:

* $h(\ldots)$ is a profile function evaluated at the daily
  temperature $\vartheta_D$.
* $KW$ is a customer-specific scaling factor in kWh/day (German: "Kundenwert").
* $F_{WT}$ is a weekday factor that adjusts for consumption differences across
  days of the week.

[^procedure]: BDEW/VKU/GEODE (2026). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Allgemeine Begriffsbestimmungen / Erläuterungen, p. VIII


### Weekday factors $F_{WT}$

Weekday factors are scaling factors applied for every day from Monday to Sunday.[^fwt]

[^holiday_rule]: BDEW/VKU/GEODE (2026). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2026-03-27. Berlin, Annex 3 (Kalender für Feiertage), p. 135.

[^fwt]: BDEW/VKU/GEODE (2026). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2026-03-27. Berlin, Annex 6, pp. 146–173 (per-profile data sheets).

For the residential profiles `HEF`, `HMF`, and `HKO`, however, all weekday factors
are equal to 1, that is, gas consumption in households is assumed *not* to differ
significantly by day of the week. 
Commercial profiles on the other hand show visible weekday patterns, the range of
the weekday factors is `r do.call(rbind, standardlastprofile:::.gas_weekday_factors) |> range() |> paste(collapse = " - ")`.

The weekday factor values are returned by `slp_gas_weekday_factors()` and
illustrated by the chart below:

```{r weekday_factors_data, echo=FALSE}
days <- c("Mo", "Tu", "We", "Th", "Fr", "Sa", "Su")
profiles_f <- names(standardlastprofile:::.gas_weekday_factors)

fwt <- do.call(rbind, lapply(profiles_f, \(pid) {
  wfct <- standardlastprofile:::.gas_weekday_factors[[pid]]
  data.frame(
    profile_id = pid,
    day        = factor(names(wfct), levels = days),
    fct_value  = unname(wfct),
    category   = if(startsWith(pid, "H")) "household" else "commercial"
  )
}))
```

```{r weekday_factors_heatmap_static, echo=FALSE, fig.asp=1, fig.retina=2, eval=!knitr::is_html_output()}
#| fig.alt = "Heatmap of gas weekday factors with one row per profile and one
#|  column per day of the week, grouped into customer categories. Colour shows
#|  the percentage deviation from the weekly mean: commercial profiles are lower
#|  at weekends and higher on weekdays, while household profiles stay close to
#|  flat across all seven days."
ggplot(fwt, aes(day, profile_id, fill = fct_value - 1)) +
  geom_tile(width = .95, height = .95) +
  coord_equal(ratio = .75) +
  scale_fill_gradient2(NULL, labels = scales::label_percent(), midpoint = 0, low = "#311dfc", mid = "#fffafe", high = "#ff005e") +
  facet_grid(rows = vars(category), scales = "free", space = "free") +
  labs(
    title   = NULL,
    caption = NULL,
    x       = NULL,
    y       = NULL,
    fill    = NULL
  ) +
  theme_minimal() +
  theme(legend.position    = "bottom") +
  theme(legend.key.width = unit(0.099,"npc")) +
  theme(panel.grid = element_blank())
```

```{r weekday_factors_heatmap_html, echo=FALSE, results='asis', eval=knitr::is_html_output()}
fwt_json <- jsonlite::toJSON(
  data.frame(
    profile_id = fwt$profile_id,
    day        = as.character(fwt$day),
    fct_value  = fwt$fct_value,
    category   = fwt$category
  ),
  dataframe = "rows"
)
# Profile descriptions for y-axis tooltips
desc_pids <- unique(as.character(fwt$profile_id))
desc_map <- vapply(desc_pids, \(pid) {
  slp_info(pid)[[pid]]$description
}, character(1))
desc_json <- jsonlite::toJSON(as.list(desc_map), auto_unbox = TRUE)
cat(sprintf('
<div id="fwt-widget" style="max-width: 600px; margin: 1.5em auto; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div id="fwt-chart"></div>
</div>
<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const data = %s;
const descriptions = %s;
const days = ["Mo","Tu","We","Th","Fr","Sa","Su"];
const householdIds = [...new Set(data.filter(d => d.category === "household").map(d => d.profile_id))];
const commerceIds  = [...new Set(data.filter(d => d.category === "commercial").map(d => d.profile_id))];
const profileOrder = [...householdIds, ...commerceIds];

const chart = Plot.plot({
  marginLeft: 40,
  marginBottom: 30,
  height: profileOrder.length * 32 + 60,
  x: { domain: days, padding: 0.08, label: null },
  y: { domain: profileOrder, padding: 0.08, label: null },
  color: {
    type: "diverging",
    domain: [-0.6, 0.3],
    pivot: 0,
    range: ["#311dfc", "#fffafe", "#ff005e"],
    label: "Weekday factors for SLP gas profiles as deviation from 1",
    tickFormat: "+.0%%",
    legend: true
  },
  marks: [
    Plot.cell(data, {
      x: "day",
      y: "profile_id",
      fill: d => d.fct_value - 1,
      tip: true,
      title: d => {
        const v = d.fct_value - 1;
        return v === 0 ? "0%%" : (v > 0 ? "+" : "") + (v * 100).toFixed(1) + "%%";
      }
    }),
    Plot.ruleY([householdIds.length - 0.5], { stroke: "#999", strokeDasharray: "3,3" })
  ]
});

document.getElementById("fwt-chart").appendChild(chart);

// Add tooltips to y-axis tick labels
chart.querySelectorAll("text").forEach(el => {
  const pid = el.textContent.trim();
  if (!descriptions[pid]) return;
  el.style.cursor = "help";
  const t = document.createElementNS("http://www.w3.org/2000/svg", "title");
  t.textContent = descriptions[pid];
  el.prepend(t);
});
</script>
', fwt_json, desc_json))
```

Public holidays are treated as a Sunday.[^holiday_rule]
24 December and 31 December are treated as Saturday unless they fall
on a Sunday; this convention mirrors the established rule from the BDEW
electricity SLP procedure and is applied here by analogy.

### Profile function $h(\vartheta_D)$

Gas consumption is strongly correlated with outdoor temperature; the colder
it gets, the more gas is consumed — at least when gas is used for space heating.
The relationship between outside temperature and gas consumption is captured by
the profile function $h(\vartheta)$:

$$
h(\vartheta) = 
\underbrace{\frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D}_{\text{sigmoid part}}
+
\underbrace{\max\!\left(m_H \vartheta + b_H,\; m_W \vartheta + b_W\right)}_{\text{linear part}}
$$


This function was initially published in 2003 as an S-shaped sigmoid function:[^Hellwig2003] 

$$
h(\vartheta) = \frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D
$$

Since then, it has been revised as part of an ongoing debate. In a
research paper from 2015 it was found that a pure sigmoid approach
systematically underestimated consumption at very cold temperatures. The authors
proposed an improved formulation called **SigLinDe**
(**Sig**moid + **Lin**ear + **De**utschland) and published SigLinDe coeffeciants
for 15 different customer groups, i.e. profile IDs.[^ffe2015] SigLinDe adds
a linear component to that sigmoid equation representing space-heating demand
(German: *Heizgas-Gerade*) and hot water demand (German: *Warmwasser-Gerade*).

The SigLinDe approach was adopted by BDEW and has since been the binding standard.[^bdew2018]

> `slp_gas()` implements the SigLinDe approach. 

### Profile IDs

The 15 gas profile IDs defined, split into residential and
commercial / industrial types.

#### Residential profiles

Residential profiles share two important characteristics: all weekday factors
equal 1 (see above), and heating load is the dominant driver of demand variability.

| Profile | Description | Notes |
|---------|-------------|-------|
| `HEF`   | Single-family home (Einfamilienhaus) | Strongest seasonal swing; includes space heating + hot water |
| `HMF`   | Multi-family home (Mehrfamilienhaus) | Lower per-unit demand than `HEF` due to shared building envelope |
| `HKO`   | Cooking and hot water only (Kochen / Warmwasser) | No space heating; TUM profile — sigmoid only, linear part is zero |

`HKO` is appropriate for customers whose gas connection serves only a cooker
or instantaneous water heater. Its flat, temperature-insensitive shape is
visible in the SigLinDe curve plot above.

#### Commercial and industrial profiles

Commercial profiles have profile-specific weekday factors that reflect the
operating hours and schedules of each sector. They are grouped here by
consumption pattern:

**Process-heat dominated** (relatively flat seasonal pattern, strong weekday
structure):

| Profile | Description |
|---------|-------------|
| `GWA`   | Laundries — large hot-water demand, insensitive to outdoor temperature |
| `GBA`   | Bakeries — continuous baking process, pronounced weekday factors |
| `GPD`   | Paper and printing — industrial process heat |

**Space-heating dominated** (strong winter peak, moderate weekday variation):

| Profile | Description |
|---------|-------------|
| `GKO`   | Small commercial (Kleinstgewerbe) |
| `GHA`   | Trade and commerce (Handel) |
| `GBD`   | Services (Dienstleistung) |
| `GBH`   | Accommodation (Beherbergung) |
| `GGA`   | Gastronomy (Gastronomie) |
| `GMK`   | Metal and automotive (Metall / Kfz) |
| `GGB`   | Mixed commercial (gemischtes Gewerbe) |
| `GMF`   | Large multi-family / mixed use (Mehrfamilienhaus groß) |

**Aggregate / summary profile**:

| Profile | Description |
|---------|-------------|
| `GHD`   | Trade, commerce and services aggregate (GHD-Stützpunkt) — weighted mean across all G-types |

`GHD` is used when a customer cannot be assigned to a specific sector. Its
parameters are a weighted average of the individual commercial profiles and
its weekday factors are close to but not equal to 1.

All 15 gas profile IDs share the same SigLinDe function — but they differ 
in their parameter values. The parameters $A$, $B$, $C$, $D$, $\vartheta_0$,
$m_H$, $b_H$, $m_W$, $b_W$ are profile-specific constants.[^params_source]

You find the complete set in the
[SigLinDe Parameters and Weekday Factors](slp-gas-parameters.html) article, and
those values are returned by `slp_gas_coefficients()`.

[^params_source]: The coefficients A, B, C, D, m~H~, b~H~, m~W~, b~W~ for each consumption profile are published in Annex 6 of the BDEW Leitfaden (pp. 146–173).

```{r siglinde-toggle-chart, echo=FALSE, results='asis', eval=knitr::is_html_output()}
.co   <- slp_gas_coefficients()
.ids  <- slp_gas_coefficients(variant = "34")$profile_id
.cols <- c("A", "B", "C", "D", "theta0", "mH", "bH", "mW", "bW")
.nest <- setNames(lapply(.ids, function(id)
  setNames(lapply(c("33", "34"), function(v)
    as.list(.co[.co$profile_id == id & .co$variant == v, .cols][1, ])),
    c("33", "34"))), .ids)
.coefs_json <- jsonlite::toJSON(.nest, auto_unbox = TRUE, digits = NA)
.descs <- vapply(.ids, function(p) slp_info(p)[[p]]$description, character(1))
.opts <- paste0(
  '<option value="', .ids, '" data-name="', .descs, '"',
  ifelse(.ids == "HEF", " selected", ""), '>', .ids, '</option>',
  collapse = ""
)
.name0 <- .descs[.ids == "HEF"]
cat(paste0('
<div id="siglinde-widget" style="width: 100%; margin: 1.5em 0; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div style="width: 80%; margin: 0 auto 12px; padding-right: 20px; box-sizing: border-box;">
<div style="font-size: 15px; font-weight: 600; color: #1a1a1a;">Higher h(\u03d1) means more gas consumption on that day</div>
<style>
.sig-toggle-label {
  display: inline-block;
  position: relative;
}
.sig-toggle-label > span {
  position: relative;
}
.sig-toggle-label::before {
  content: attr(data-text);
  font-weight: 700;
  visibility: hidden;
  display: block;
  height: 0;
  overflow: hidden;
  pointer-events: none;
}
.sig-active {
  font-weight: 700;
  color: #000;
  text-decoration: underline;
  text-decoration-thickness: 1px;
  text-underline-offset: 2px;
}
.sig-select-box {
  display: inline-block;
  position: relative;       /* the box -- not the native select -- owns the height */
  top: -2px;                /* nudge up to centre in the text line */
  width: 4em;
  height: 1.2em;            /* same as the toggle */
  box-sizing: border-box;
  overflow: hidden;
  vertical-align: middle;
  margin: 0 0 0 2px;
  background-color: #fff;
  outline: 2px solid #333;   /* outline (outside the box) to match the toggle */
  cursor: pointer;
  background-image:
    linear-gradient(45deg, transparent 50%, #333 50%),
    linear-gradient(135deg, #333 50%, transparent 50%);
  background-position: calc(100% - 7px) center, calc(100% - 4px) center;
  background-size: 5px 5px, 5px 5px;
  background-repeat: no-repeat;
}
#siglinde-profile {
  position: absolute;       /* taller native select, vertically centred + clipped */
  top: 50%;
  left: 0;
  transform: translateY(-50%);
  width: 100%;
  box-sizing: border-box;
  appearance: none;
  -webkit-appearance: none;
  -moz-appearance: none;
  border: none;
  outline: none;
  margin: 0;
  padding: 0 16px 0 5px;
  background: transparent;
  font-family: inherit;
  font-size: 11px;
  font-weight: 700;
  line-height: 1;
  color: #1a1a1a;
  cursor: pointer;
}
</style>
<p style="font-size: 13px; color: #555; margin: 4px 0 0; line-height: 1.4;">Toggle between <span id="siglinde-label-sig" class="sig-toggle-label" data-text="Sigmoid"><span id="siglinde-label-sig-inner" class="sig-active">Sigmoid</span></span>
<span style="display: inline-block; vertical-align: middle; position: relative; top: -2px; width: 2.2em; height: 1.2em; margin: 0 2px;">
<button id="siglinde-toggle" role="switch" aria-checked="false" aria-label="Toggle between a pure Sigmoid and SigLinDe"
  style="position: relative; width: 100%; height: 100%; background: #fff; outline: 2px solid #333; border: none; border-radius: 0; cursor: pointer; padding: 0;">
<span id="siglinde-thumb" style="position: absolute; top: 50%; left: 0.15em; width: 0.75em; height: 0.75em; background: #333; border-radius: 0; display: block; transition: transform 0.2s ease; transform: translateY(-50%);"></span>
</button>
</span>
and <span id="siglinde-label-sld" class="sig-toggle-label" data-text="SigLinDe"><span id="siglinde-label-sld-inner">SigLinDe</span></span> functions for profile <span class="sig-select-box"><select id="siglinde-profile">', .opts, '</select></span></p>
<p style="font-size: 13px; color: #555; margin: 6px 0 0; line-height: 1.4; text-align: left;"><span id="siglinde-profile-name" style="font-weight: 600; color: #1a1a1a;">', .name0, '</span></p>
</div>
<div id="siglinde-chart" style="width: 80%; margin: 8px auto 0;"></div>
</div>

<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const coefs = ', .coefs_json, ';
const variants = ["33", "34"];
let p = coefs["HEF"];          // p["33"], p["34"]

function sigmoid(c, t) {
  return c.A / (1 + Math.pow(c.B / (t - c.theta0), c.C)) + c.D;
}
function linear(c, t) {
  return Math.max(c.mH * t + c.bH, c.mW * t + c.bW);
}

const temps = [];
for (let t = -15; t <= 25; t += 0.25) temps.push(t);
const i8 = temps.indexOf(8);   // index of the 8°C reference

function curvesFor(c) {
  return {
    sig: temps.map(t => sigmoid(c, t)),
    sld: temps.map(t => sigmoid(c, t) + linear(c, t))
  };
}
// displayed (possibly mid-morph) curves, one per variant
let curves = { "33": curvesFor(p["33"]), "34": curvesFor(p["34"]) };

const chartEl  = document.getElementById("siglinde-chart");
const toggle   = document.getElementById("siglinde-toggle");
const thumb    = document.getElementById("siglinde-thumb");
const labelSig = document.getElementById("siglinde-label-sig-inner");
const labelSld = document.getElementById("siglinde-label-sld-inner");

let active = true;
let blend  = 1;
let animId = null;
labelSig.classList.remove("sig-active");
labelSld.classList.add("sig-active");
toggle.setAttribute("aria-checked", "true");
thumb.style.transform = "translateY(-50%) translateX(1.1em)";

function render(t) {
  const width = chartEl.clientWidth;
  if (!width) return;
  const data = [];
  const refDots = [];
  for (const v of variants) {
    const c = curves[v];
    const label = "Variant " + v;
    temps.forEach((temp, i) => {
      data.push({ variant: label, temp, h: c.sig[i] + t * (c.sld[i] - c.sig[i]) });
    });
    refDots.push({ variant: label, temp: 8, h: c.sig[i8] + t * (c.sld[i8] - c.sig[i8]) });
  }
  const plot = Plot.plot({
    width,
    height: Math.round(width * 0.6),
    marginLeft: 44,
    marginBottom: 36,
    color: {
      domain: ["Variant 33", "Variant 34"],
      range: ["#961BFA", "#0CC792"],
      legend: true
    },
    x: { label: "Temperature (\u00b0C)", ticks: [-10, 0, 8, 20], tickFormat: d => d + "\u00b0", grid: true },
    y: { label: "h(\u03d1)", domain: [0, 6], ticks: [1, 2, 3, 4, 5], grid: true },
    marks: [
      Plot.ruleY([1], { stroke: "#999", strokeDasharray: "4,3" }),
      Plot.lineY(data, { x: "temp", y: "h", stroke: "variant", strokeWidth: 2.5 }),
      Plot.dot(refDots, { x: "temp", y: "h", fill: "variant", r: 3.5, stroke: "white", strokeWidth: 1 })
    ],
    style: {
      fontSize: "12px",
      fontFamily: "-apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif",
      background: "transparent"
    }
  });
  chartEl.replaceChildren(plot);
}

function animate() {
  const target = active ? 1 : 0;
  const step = 0.06;
  const diff = target - blend;
  if (Math.abs(diff) < step) {
    blend = target;
    render(blend);
    animId = null;
    return;
  }
  blend += Math.sign(diff) * step;
  render(blend);
  animId = requestAnimationFrame(animate);
}

toggle.addEventListener("click", () => {
  active = !active;
  toggle.setAttribute("aria-checked", active);
  thumb.style.transform = active ? "translateY(-50%) translateX(1.1em)" : "translateY(-50%)";
  labelSig.classList.toggle("sig-active", !active);
  labelSld.classList.toggle("sig-active", active);
  if (animId) cancelAnimationFrame(animId);
  animId = requestAnimationFrame(animate);
});

const profileSelect = document.getElementById("siglinde-profile");
let morphId = null;
profileSelect.addEventListener("change", () => {
  const from = {};
  for (const v of variants) {
    from[v] = { sig: curves[v].sig.slice(), sld: curves[v].sld.slice() };
  }
  p = coefs[profileSelect.value];
  document.getElementById("siglinde-profile-name").textContent =
    profileSelect.selectedOptions[0].dataset.name;
  const to = { "33": curvesFor(p["33"]), "34": curvesFor(p["34"]) };
  let m = 0;
  if (morphId) cancelAnimationFrame(morphId);
  (function step() {
    m = Math.min(1, m + 0.08);
    for (const v of variants) {
      curves[v] = {
        sig: from[v].sig.map((x, i) => x + m * (to[v].sig[i] - x)),
        sld: from[v].sld.map((x, i) => x + m * (to[v].sld[i] - x))
      };
    }
    render(blend);
    morphId = m < 1 ? requestAnimationFrame(step) : null;
  })();
});

window.addEventListener("resize", () => render(blend));

render(1);
</script>
'))
```


[^Hellwig2003]: Hellwig, Mark (2003). *Entwicklung und Anwendung parametrisierter Standard-Lastprofile*. Dissertation, TU München.

[^ffe2015]: Hinterstocker, M., Eberl, B., von Roon, S. (2015). *Weiterentwicklung des Standardlastprofilverfahrens Gas*. Endbericht. Forschungsgesellschaft für Energiewirtschaft mbH (FfE), available at <https://web.archive.org/web/20260620061251/https://www.bdew.de/media/documents/201507_Weiterentwicklung-SLP-Gas.pdf>

[^bdew2018]: BDEW/VKU/GEODE (2026). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2026-03-27. Available at <https://web.archive.org/web/20260619125016/https://www.bdew.de/media/documents/260327_LF_SLP_Gas_KoV_XV_CO4f7Rb.pdf>

The result of $h(\vartheta)$ is a number > 0 without a dimension:

``` {r h-vartheta-minimal-example, message=FALSE}
# profile "HEF", Ausprägung: 34
slp_gas_siglinde(
  theta  =   8.0,
  A      =   1.381966,
  B      = -37.41242,
  C      =   6.172318,
  D      =   0.0396284,
  theta0 =  40.0,
  mH     =  -0.0672159,
  bH     =   1.116714,
  mW     =  -0.0019982,
  bW     =   0.135507
)
```

For `theta = 8` the result above is ~1 — which is no coincidence, and brings us
to the `kundenwert`.

### Kundenwert $KW$

`kundenwert` accounts for a customer’s individual consumption behaviour. It
answers the question: how much gas does a customer consume on a day when the
outside temperature is 8°C? It is a scaling factor in kWh/day — 8°C is the
temperature at which $h(\vartheta) = 1$ for every profile, with one exception:
`HKO`, whose linear parameters ($m_H$, $b_H$, $m_W$, $b_W$) are all 0. With no
linear component, the `HKO` profile function is a plain sigmoid rather than a
SigLinDe (sigmoid-plus-linear) function, and so is not normalised to 1 at 8°C.

`kundenwert` is derived from the customer’s annual consumption and the year’s
temperature and weekday profile. Starting with $\eqref{eq:qdaily}$, the annual total
consumption is the sum of all daily values:

$$
Q_a = \sum_{D} Q(D) = \sum_{D} h(\vartheta_D) \times KW \times F_{WT,D}
$$

Rearranging to solve for $KW$:

$$
KW = \frac{Q_a}{\displaystyle\sum_{D} h(\vartheta_D) \times F_{WT,D}} \tag{2}\label{eq:kw}
$$

Where

* $Q_a$ is the annual gas consumption (German: *Jahresverbrauch*)
* $\vartheta_D$ is the daily temperature for day $D$
* $F_{WT,D}$ is the weekday factor for day $D$

Using the example of a single-family home (`HEF`) in Düsseldorf with an annual
consumption of 15,000 kWh — [the same customer as in the README](https://flrd.github.io/standardlastprofile/#gas) — the table
below shows how the denominator is built up, day by day, from the Düsseldorf
long-term daily mean temperatures (2004–2024):

```{r kundenwert_table, echo=FALSE, message=FALSE}
# Düsseldorf long-term daily mean temperature (DWD station 1078, mean over
# 2004-2024) -- the same reference series behind the Kundenwert in the README.
# Embedded here so the worked example is self-contained; see "Deriving the
# Kundenwert" below for how to fetch it.
ref_dates <- seq.Date(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day")
ref <- data.frame(
  date        = ref_dates,
  temp_c_mean = c(
    4.8, 4.2, 4.2, 4.6, 4.5, 4.0, 4.4, 4.6, 4.4, 4.2, 4.4, 4.8, 4.8,
    4.2, 3.6, 3.5, 3.5, 3.3, 3.1, 3.1, 2.5, 2.4, 2.2, 2.4, 2.1, 2.4,
    2.9, 3.3, 3.6, 3.4, 3.5, 4.1, 4.2, 4.6, 4.6, 4.7, 4.5, 3.1, 3.0,
    3.3, 3.4, 3.4, 2.9, 3.1, 4.2, 5.0, 5.6, 5.3, 4.9, 4.5, 5.1, 5.4,
    5.6, 5.2, 5.3, 5.0, 4.7, 3.9, 4.0, 4.6, 5.0, 4.9, 5.1, 4.5, 4.7,
    4.8, 5.7, 6.6, 6.6, 6.2, 6.3, 6.9, 6.8, 7.6, 8.3, 7.8, 7.2, 7.3,
    7.0, 7.0, 7.3, 7.2, 7.5, 7.7, 8.0, 8.3, 8.8, 8.8, 9.9, 9.9, 9.1,
    9.4, 9.6, 8.6, 8.1, 9.6, 10.2, 9.4, 9.7, 10.2, 10.0, 10.4, 10.2,
    10.6, 10.7, 10.9, 10.6, 10.9, 11.0, 11.3, 11.6, 11.5, 11.7, 12.3,
    12.3, 11.6, 11.5, 12.0, 12.3, 12.5, 12.5, 12.4, 12.1, 12.2, 12.2,
    13.1, 13.5, 14.1, 15.0, 14.9, 14.3, 13.7, 13.5, 13.5, 13.0, 13.3,
    13.1, 14.3, 14.5, 15.2, 16.0, 15.8, 15.1, 15.0, 15.3, 15.8, 16.0,
    16.2, 15.6, 15.4, 15.6, 15.9, 17.1, 17.1, 16.9, 16.7, 16.7, 17.0,
    17.6, 17.7, 17.9, 17.8, 16.9, 16.8, 17.2, 17.6, 17.0, 18.3, 19.2,
    18.4, 18.1, 17.6, 17.6, 18.3, 19.0, 19.1, 18.7, 19.0, 19.6, 19.4,
    19.2, 18.9, 19.2, 19.2, 19.5, 19.2, 18.7, 18.7, 18.6, 19.0, 19.1,
    18.9, 18.4, 18.4, 18.4, 19.2, 19.8, 20.0, 20.7, 21.1, 20.3, 19.8,
    20.1, 20.0, 20.1, 20.3, 20.4, 20.5, 20.0, 19.7, 19.6, 19.3, 20.1,
    19.7, 19.9, 19.6, 19.9, 20.3, 20.0, 18.8, 18.9, 18.5, 18.8, 19.1,
    19.0, 19.0, 19.3, 19.0, 18.6, 19.1, 19.3, 19.0, 18.7, 18.6, 19.0,
    19.1, 18.4, 18.5, 17.8, 17.9, 17.4, 17.2, 17.5, 16.8, 17.3, 17.5,
    17.7, 18.1, 17.9, 17.4, 17.2, 17.2, 17.3, 17.2, 16.6, 16.3, 15.8,
    15.8, 15.6, 15.0, 14.7, 14.4, 14.7, 15.0, 14.7, 14.8, 14.9, 14.1,
    14.1, 13.9, 13.9, 13.7, 13.4, 13.7, 13.4, 13.4, 13.3, 13.4, 13.3,
    13.2, 12.5, 12.1, 12.5, 12.0, 11.8, 12.3, 11.5, 11.3, 11.3, 11.5,
    11.4, 12.0, 11.9, 12.0, 11.3, 11.1, 11.9, 11.9, 12.1, 11.3, 10.9,
    11.0, 10.5, 10.3, 10.5, 10.7, 9.9, 9.7, 9.0, 8.9, 9.0, 8.8, 8.7,
    8.2, 8.0, 7.6, 8.0, 7.8, 7.9, 7.5, 7.3, 7.1, 6.8, 6.2, 6.0, 6.2,
    6.5, 6.0, 6.0, 5.2, 5.6, 4.9, 4.4, 4.0, 4.5, 4.3, 3.9, 4.1, 4.8,
    5.3, 5.6, 5.1, 4.5, 3.8, 4.0, 4.1, 3.5, 3.8, 4.4, 4.6, 4.7, 4.4,
    4.6, 3.9, 4.9, 6.3, 6.7, 6.5, 5.3, 4.5, 3.8, 3.7, 3.9, 4.7, 5.7
  )
)

# Compute h(theta) for HEF, variant 34
p <- standardlastprofile:::.gas_profile_params[["34"]][["HEF"]]
ref$h <- slp_gas_siglinde(
  ref$temp_c_mean, p$A, p$B, p$C, p$D, p$theta0,
  p$mH, p$bH, p$mW, p$bW
)

# Weekday factors (all 1 for HEF, but include for completeness)
wt_keys <- standardlastprofile:::.get_gas_weekday_key(ref$date)
fwt <- standardlastprofile:::.gas_weekday_factors[["HEF"]]
ref$f_wt <- unname(fwt[wt_keys])

# Show first and last few rows with ellipsis
show <- rbind(
  head(ref, 1),
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  ref[floor(nrow(ref) / 2) + c(-1, 0, 1), ],
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  tail(ref, 1)
)
show_df <- data.frame(
  Date = ifelse(is.na(show$date), "\u22ee", format(show$date, "%b %e")),
  `°C` = ifelse(is.na(show$temp_c_mean), "\u22ee", round(show$temp_c_mean, 1)),
  `h(ϑ)` = ifelse(is.na(show$h), "\u22ee", round(show$h, 4)),
  check.names = FALSE
)
knitr::kable(show_df, align = "lrr")
```

The sum in the denominator across all `r nrow(ref)` days gives
$\sum_D h(\vartheta_D) = 272.32$. And since for all three residential profiles the
weekday factors are 1, $\eqref{eq:kw}$ simplifies to:

$$
KW = \frac{Q_a}{\displaystyle\sum_{D} h(\vartheta_D)}
$$

So $KW \approx$ 15,000 / `r round(sum(ref$h * ref$f_wt), 2)` =
**`r round(15000 / sum(ref$h * ref$f_wt), 1)`** kWh/day.

## A working example: Düsseldorf

With the algorithm covered, let us see how to use `slp_gas()` with real-world
temperature data.

`slp_gas()` takes:

- a `profile_id`
- a `dates` vector, and a matching vector of `temperatures`, and 
- a `kundenwert`.

### Deriving the Kundenwert

To derive `kundenwert` we need an annual consumption and temperature data. We
can feed this information into the function `slp_gas_kundenwert()` together
with any of the 15 gas profile IDs. To produce a meaningful result the
temperature series should cover at least a full year. We show you how to fetch
measured data from a trustworthy institution like the German Meteorological
Service, or DWD for short.

The [`rdwd`](https://cran.r-project.org/package=rdwd) package provides a
straightforward interface to the open-data portal of the DWD 
(Deutscher Wetterdienst).[^dwd] No registration or API key is required.

[^dwd]: DWD Open Data: <https://opendata.dwd.de/>

#### rdwd package 

This section gives you a quick introduction to the `rdwd` package, read more
about its features at [brry.github.io/rdwd](https://brry.github.io/rdwd/index.html).

Our workflow has three steps:

1. `selectDWD()` — find the download URL for a station and dataset.
2. `dataDWD()` — download and unzip the file.
3. `readDWD()` — parse the file into a data frame.

We need the daily climate summary (`var = "kl"`, `res = "daily"`). The
column `TMK` contains the daily mean temperature in °C.
To find the exact station name to pass to `selectDWD()`, use `findID()`:

```{r dwd_stations, eval=FALSE}
# returns matching station names and IDs
rdwd::findID("Duesseldorf", exactmatch = FALSE) 
```

For the reference temperatures we use the long-term daily mean
over 2004–2024 rather than a single year, so that no individual-year anomaly
distorts the scaling factor^[BDEW Leitfaden, section
3.6.3]. 

```{r fetch-dwd, eval=FALSE}
library(rdwd)

# Full daily climate record for Düsseldorf (DWD station 1078)
links <- selectDWD(
  "Duesseldorf",
  res = "daily",
  var = "kl",
  per = c("historical", "recent"),
  current = TRUE
)

raw <- do.call(
  rbind,
  readDWD(dataDWD(links, read = FALSE, force = TRUE), varnames = FALSE)
)

# drop historical/recent overlap
raw <- raw[!duplicated(raw$MESS_DATUM), ]   

# Long-term mean daily temperature over 2004-2024, one value per calendar day
ref       <- raw[format(raw$MESS_DATUM, "%Y") %in% as.character(2004:2024), ]
ref$mmdd  <- format(ref$MESS_DATUM, "%m-%d")
clim      <- aggregate(TMK ~ mmdd, data = ref, FUN = mean)
clim      <- clim[clim$mmdd != "02-29", ]          # drop the leap day
dates_ref <- as.Date(paste0("2023-", clim$mmdd))   # any non-leap year
temps_ref <- clim$TMK
```

Now that we have a reference temperature series, we can call
`slp_gas_kundenwert()`:

```{r prep_kw, include=FALSE}
dates_ref <- ref_dates
temps_ref <- ref$temp_c_mean
```

```{r kw_derive, eval=TRUE}
# kundenwert for a 15,000 kWh/a HEF customer
slp_gas_kundenwert("HEF", dates_ref, temps_ref, annual_consumption = 15000)
```

That value is what the README example passes to `slp_gas()` to generate the
2025/26 heating season — and it is what we use below to ask how the same
customer would fare in a different climate. 

### Same customer preferences, different climate

We keep that `kundenwert` of 55.1 kWh/day fixed and ask how the gas consumption
changes in another place — same customer, same KW, different
climate. Germany spans several distinct climate zones; we compare three:

* **Chemnitz** — continental climate in the Ore Mountains foothills of Saxony;
  colder winters and the greatest temperature amplitude of the three.
* **Freiburg im Breisgau** — mild oceanic-continental climate in the sheltered
  Upper Rhine Plain; the warmest, sunniest large city in Germany.
* **Hamburg** — maritime climate on the North Sea coast; mild but cloudy,
  windy winters.

The grid below has one column per comparison city and one row per month of the
2025/26 heating season (the same period as the README example). Each point is a
day, plotting the city's daily gas consumption (y-axis) against Düsseldorf
(x-axis); the 45° line marks equal consumption.

That winter all three cities ran colder than Düsseldorf, so every point cloud
sits above the line — the question is by how much:

- **Chemnitz** is highest, about **35 % more gas** over the season
  (≈ 16,400 vs 12,200 kWh);
- **Freiburg im Breisgau** is closest to Düsseldorf at about **11 % more**;
- **Hamburg** follows at about **25 % more**.

The gap is widest on the cold January and February days at the top-right of each
panel, and narrows in the mild shoulder months of October and April where the
clouds approach the line.

```{r gas_cities_chart, echo=FALSE, results='asis', eval=knitr::is_html_output()}
cities_df <- standardlastprofile:::.gas_cities_data()
cities_json <- jsonlite::toJSON(
  data.frame(
    city        = as.character(cities_df$city),
    month       = as.character(cities_df$month),
    # month.abb is always English regardless of locale (avoids "Okt" etc.)
    label       = paste(
      as.integer(format(cities_df$date, "%d")),
      month.abb[as.integer(format(cities_df$date, "%m"))]
    ),
    duesseldorf = round(cities_df$duesseldorf, 1),
    other       = round(cities_df$other, 1)
  ),
  dataframe = "rows"
)
cat(paste0('
<div id="cities-widget" style="width: 100%; margin: 1.5em 0; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div style="font-size: 15px; font-weight: 600; color: #1a1a1a;">SLP Gas -- single family home, three climates vs Düsseldorf</div>
<p style="font-size: 13px; color: #555; margin: 4px 0 12px; line-height: 1.4;">Each point is one heating-season day. Points above the 45° line used more gas than in Düsseldorf.</p>
<style>
#cities-chart { display: grid; grid-template-columns: repeat(auto-fill, minmax(200px, 1fr)); gap: 10px; align-items: start; }
#cities-chart .city-h { text-align: center; font-weight: 700; font-size: 13px; color: #1a1a1a; }
</style>
<div id="cities-chart" role="group" aria-label="Daily gas consumption in Chemnitz, Freiburg im Breisgau and Hamburg plotted against Düsseldorf for each month of the heating season, with a 45-degree reference line. All three cities sit above the line that winter, furthest for Chemnitz, then Hamburg, with Freiburg im Breisgau closest.">
<div><div class="city-h">Chemnitz</div><div class="city-plot" data-city="Chemnitz"></div></div>
<div><div class="city-h">Freiburg im Breisgau</div><div class="city-plot" data-city="Freiburg"></div></div>
<div><div class="city-h">Hamburg</div><div class="city-plot" data-city="Hamburg"></div></div>
</div>
</div>
<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const data = ', cities_json, ';
const cities = ["Chemnitz", "Freiburg", "Hamburg"];
const months = ["Oct", "Nov", "Dec", "Jan", "Feb", "Mar", "Apr"];
const hi  = Math.max(...data.flatMap(d => [d.duesseldorf, d.other]));
const lim = [0, Math.ceil(hi / 20) * 20];

const panels = cities.map(city => ({
  city,
  el: document.querySelector(`#cities-chart .city-plot[data-city="${city}"]`),
  rows: data.filter(d => d.city === city)
}));

// uniform margins on every plot (y-axis reserved left, month labels right) so
// the cells stay identical squares whether or not a given plot draws them
const mL = 44, mR = 34, mT = 8, mB = 34, fyPad = 0.06;

function plotCity(panel, width, showY, showMonths) {
  // x and y share the kWh scale, so each month cell must be exactly square (no
  // stretch). The 7 facet rows each lose fyPad of their vertical step to
  // padding, so divide it back out so the drawn cell height equals its width
  // at any column count.
  const cellW = width - mL - mR;
  const height = Math.round((7 * cellW) / (1 - fyPad) + mT + mB);
  const marks = [
    Plot.line(panel.rows, {
      fy: "month",
      x: "duesseldorf", y: "duesseldorf",
      stroke: "#333", strokeWidth: 0.6, sort: "duesseldorf"
    }),
    Plot.dot(panel.rows, {
      fy: "month",
      x: "duesseldorf", y: "other",
      fill: "#0CC792", r: 1.8, fillOpacity: 0.85,
      tip: true,
      title: d => d.city + ": " + d.other.toFixed(1) + " kWh\\n" +
                  "Düsseldorf: " + d.duesseldorf.toFixed(1) + " kWh\\n" + d.label
    })
  ];
  // month labels on the right, drawn only on the rightmost plot of each row
  if (showMonths) {
    marks.push(Plot.axisFy({ anchor: "right", label: null, tickSize: 0 }));
  }
  return Plot.plot({
    width,
    height,
    marginLeft: mL,
    marginRight: mR,
    marginTop: mT,
    marginBottom: mB,
    fy: { domain: months, axis: null, padding: fyPad },
    // x-axis ticks on every plot, but the title only on the rightmost of each row
    x: { domain: lim, label: showMonths ? "Düsseldorf (kWh/day)" : null, ticks: 3, grid: true },
    // y-axis (ticks + title) drawn only on the leftmost plot of each row
    y: showY
      ? { domain: lim, label: "Comparison city (kWh/day)", ticks: 3, grid: true }
      : { domain: lim, axis: null, grid: true },
    style: {
      fontSize: "11px",
      fontFamily: "-apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif",
      background: "transparent"
    },
    marks
  });
}

function render() {
  // label each visual row once: the leftmost plot gets the y-axis, the rightmost
  // gets the month labels -- recomputed on every (re)render so it follows the
  // 3- / 2- / 1-up reflow.
  const rects = panels.map(p => p.el.getBoundingClientRect());
  const minLeft = Math.min(...rects.map(r => r.left));
  panels.forEach((panel, i) => {
    const width = panel.el.clientWidth;
    if (!width) return;
    const r = rects[i];
    const firstInRow = Math.abs(r.left - minLeft) < 1;
    const lastInRow = !rects.some(
      (o, j) => j !== i && Math.abs(o.top - r.top) < 1 && o.left > r.left + 1
    );
    const chart = plotCity(panel, width, firstInRow, lastInRow);
    chart.querySelectorAll(`[aria-label="fy-axis tick label"] text`)
      .forEach(t => { t.style.fontWeight = "700"; });
    panel.el.replaceChildren(chart);
  });
}

render();
window.addEventListener("resize", render);
</script>
'))
```
