---
title: "Generate a standard load profile for gas"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  dpi       = 300,
  fig.asp   = 0.618,
  fig.width = 6,
  fig.height = 3,
  fig.align = "center",
  out.width = "90%"
)
```

```{=html}
<style>
  .content h3 {
    margin-top: -30px !important;
  }

  details {
    margin-bottom: 40px;
  }

  table {
    width: 100%;
    table-layout: auto;
  }

  table col {
    width: auto !important;
  }
</style>
```

``` {r, message=FALSE, include=FALSE}
library(standardlastprofile)
library(ggplot2)
```

A standard load profile reflects the typical pattern of gas consumption. It
is used for forecasting and balancing purposes for various customer types or
customer groups for which no continuous measurement data is available. In this
respect, standard load profiles represent a simplification. 

In this article we explain how `slp_gas()` implements the algorithm of the
**synthetic procedure** (German: *synthetisches Verfahren*), following the
definitions by BDEW, VKU, and GEODE in
the *Leitfaden Abwicklung von Standardlastprofilen Gas* (2025).[^leitfaden]

[^leitfaden]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin. Available at <https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>

## Synthetic procedure

The synthetic procedure takes a bottom-up approach: the gas consumption for a
gas network is explained by the sum of the individual gas consumers within that
network.^[In addition to the synthetic procedure, there is also the analytical
procedure. In brief, under this procedure, a network operator forecasts
the proportion of the daily gas volume attributable to *all* SLP customers
based on the expected temperature for that day and the residual gas amount
derived from the total gas consumption, minus the gas volume attributable to
customers with metered consumption. The analytical procedure is not in scope of
this package.]

For each customer type, i.e. profile, the gas
consumption of day $D$ is the result of three factors:

1. outside temperature
2. preferences and habits
3. the day of the week

Mathematically, this is expressed as:[^procedure]

$$
Q(D) = h(\vartheta_D) \times KW \times F_{WT, D}
$$
where:

* $h(\ldots)$ is a profile function evaluated at the daily
  temperature $\vartheta_D$.
* $KW$ is a customer-specific scaling factor in kWh/day (German: "Kundenwert").
* $F_{WT}$ is a weekday factor that adjusts for consumption differences across
  days of the week.

[^procedure]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*. Berlin, Allgemeine Begriffsbestimmungen / Erläuterungen, p. VIII

### Profile function $h(\vartheta_D)$

Gas consumption is strongly correlated with outdoor temperature; the colder
it gets, the more gas is consumed. The relationship between outside
temperature and gas consumption is captured by the profile function $h(\vartheta)$. 

This function was initially published in 2003 as an S-shaped sigmoid function:[^Hellwig2003] 

$$
h(\vartheta) = \frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D
$$

Since then, it has been revised as part of an ongoing debate. In a
research paper from 2015 it was found that a pure sigmoid approach
systematically underestimated consumption at very cold temperatures. The authors
proposed an improved formulation called **SigLinDe**
(**Sig**moid + **Lin**ear + **De**utschland).[^ffe2015] SigLinDe adds
a linear component to that sigmoid equation representing space-heating demand
(German: *Heizgas-Gerade*) and hot water demand (German: *Warmwasser-Gerade*).

$$
h(\vartheta) = 
\underbrace{\frac{A}{1 + \left(\dfrac{B}{\vartheta - \vartheta_0}\right)^C} + D}_{\text{sigmoid part}}
+
\underbrace{\max\!\left(m_H \vartheta + b_H,\; m_W \vartheta + b_W\right)}_{\text{linear part}}
$$

```{r siglinde-toggle-chart, echo=FALSE, results='asis', eval=knitr::is_html_output()}
cat('
<div id="siglinde-widget" style="max-width: 600px; margin: 1.5em auto; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div style="margin-bottom: 12px;">
<div style="font-size: 15px; font-weight: 600; color: #1a1a1a;">Higher h(\u03d1) means more gas consumption on that day</div>
<style>
.sig-toggle-label {
  display: inline-block;
  position: relative;
}
.sig-toggle-label > span {
  position: relative;
}
.sig-toggle-label::before {
  content: attr(data-text);
  font-weight: 700;
  visibility: hidden;
  display: block;
  height: 0;
  overflow: hidden;
  pointer-events: none;
}
</style>
<p style="font-size: 13px; color: #555; margin: 4px 0 0; line-height: 1.4;">Toggle between <span id="siglinde-label-sig" class="sig-toggle-label" data-text="Sigmoid"><span id="siglinde-label-sig-inner" style="font-weight: 700; color: #0CC792;">Sigmoid</span></span>
<span style="display: inline-block; vertical-align: middle; width: 2.2em; height: 1.2em; margin: 0 2px;">
<button id="siglinde-toggle" role="switch" aria-checked="false" aria-label="Toggle between a pure Sigmoid and SigLinDe"
  style="position: relative; width: 100%; height: 100%; background: #fff; outline: 2px solid #333; border: none; border-radius: 0; cursor: pointer; padding: 0;">
<span id="siglinde-thumb" style="position: absolute; top: 50%; left: 0.15em; width: 0.75em; height: 0.75em; background: #333; border-radius: 0; display: block; transition: transform 0.2s ease; transform: translateY(-50%);"></span>
</button>
</span>
and <span id="siglinde-label-sld" class="sig-toggle-label" data-text="SigLinDe"><span id="siglinde-label-sld-inner" style="font-weight: 400;">SigLinDe</span></span> functions</p>
</div>
<div id="siglinde-chart"></div>
</div>

<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const A = 1.0443538, B = -35.0333754, C = 6.2240634, D = 0.0502917, theta0 = 40;
const mH = -0.053583, bH = 0.9995901, mW = -0.0021758, bW = 0.1633299;

function sigmoid(t) {
  return A / (1 + Math.pow(B / (t - theta0), C)) + D;
}
function linear(t) {
  return Math.max(mH * t + bH, mW * t + bW);
}

const temps = [];
for (let t = -15; t <= 25; t += 0.25) temps.push(t);

const sigOnly  = temps.map(t => sigmoid(t));
const siglinde = temps.map(t => sigmoid(t) + linear(t));

const chartEl  = document.getElementById("siglinde-chart");
const toggle   = document.getElementById("siglinde-toggle");
const thumb    = document.getElementById("siglinde-thumb");
const labelSig = document.getElementById("siglinde-label-sig-inner");
const labelSld = document.getElementById("siglinde-label-sld-inner");

let active = false;
let blend  = 0;
let animId = null;

function render(t) {
  const data = temps.map((temp, i) => ({
    temp,
    h: sigOnly[i] + t * (siglinde[i] - sigOnly[i])
  }));

  const marks = [
    Plot.lineY(data, {
      x: "temp", y: "h",
      stroke: "#0CC792", strokeWidth: 2.5
    })
  ];

  const plot = Plot.plot({
    width: 600,
    height: 340,
    marginLeft: 50,
    marginBottom: 40,
    x: {
      label: "Temperature (\u00b0C)",
      tickFormat: d => d + "\u00b0",
      grid: true
    },
    y: {
      label: "h(\u03d1)",
      domain: [0, 3],
      grid: true
    },
    marks,
    style: {
      fontSize: "13px",
      fontFamily: "-apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif",
      background: "transparent"
    }
  });

  chartEl.innerHTML = "";
  chartEl.appendChild(plot);
}

function animate() {
  const target = active ? 1 : 0;
  const step = 0.06;
  const diff = target - blend;
  if (Math.abs(diff) < step) {
    blend = target;
    render(blend);
    animId = null;
    return;
  }
  blend += Math.sign(diff) * step;
  render(blend);
  animId = requestAnimationFrame(animate);
}

toggle.addEventListener("click", () => {
  active = !active;
  toggle.setAttribute("aria-checked", active);
  thumb.style.transform = active ? "translateY(-50%) translateX(1.1em)" : "translateY(-50%)";
  labelSig.style.fontWeight = active ? "400" : "700";
  labelSig.style.color = active ? "" : "#0CC792";
  labelSld.style.fontWeight = active ? "700" : "400";
  labelSld.style.color = active ? "#0CC792" : "";
  if (animId) cancelAnimationFrame(animId);
  animId = requestAnimationFrame(animate);
});

render(0);
</script>
')
```

The SigLinDe approach was adopted by BDEW and has since been the binding standard.[^bdew2018]
`slp_gas()` implements the SigLinDe approach. The parameters $A$,
$B$, $C$, $D$, $\vartheta_0$, $m_H$, $b_H$, $m_W$, $b_W$ are profile-specific
constants.[^params_source]

All **15 gas profile IDs** share this one SigLinDe function — they differ only
in their parameter values ($A$, $B$, …, $b_W$). The complete set is tabulated
in the
[SigLinDe Parameters and Weekday Factors](slp-gas-parameters.html) article and
returned in R by `slp_gas_coefficients()`:

```{r coef_example, message=FALSE}
slp_gas_coefficients("HEF")
```

The `HKO` profile (cooking and hot water only) is a special case: all four
linear coefficients ($m_H$, $b_H$, $m_W$, $b_W$) are zero, so the linear term
$\max(m_H\vartheta + b_H,\; m_W\vartheta + b_W)$ vanishes and $h(\vartheta)$
collapses to the bare sigmoid — it is the original pre-SigLinDe TU München
profile, carried over unchanged. Because the variant 33 / 34 split only affects
the linear part, the two variants are **identical for `HKO`**.

[^params_source]: The coefficients A, B, C, D, m~H~, b~H~, m~W~, b~W~ for each consumption profile are published in Annex 6 of the BDEW Leitfaden (pp. 140–163).

[^Hellwig2003]: Hellwig, Mark (2003). *Entwicklung und Anwendung parametrisierter Standard-Lastprofile*. Dissertation, TU München.

[^ffe2015]: Hinterstocker, M., Eberl, B., von Roon, S. (2015). *Weiterentwicklung des Standardlastprofilverfahrens Gas*. Endbericht. Forschungsgesellschaft für Energiewirtschaft mbH (FfE), available at <https://www.bdew.de/media/documents/201507_Weiterentwicklung-SLP-Gas.pdf>

[^bdew2018]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2025-10-28. Available at <https://www.bdew.de/media/documents/251028_LF_SLP_Gas_KoV_XIV.2.pdf>

The result of $h(\vartheta)$ is a number > 0 without a dimension:

``` {r h-vartheta-minimal-example, message=FALSE}
# profile "GMF", Ausprägung: 34
slp_gas_siglinde(
  theta  =   8.0,
  A      =   1.0443538,
  B      = -35.0333754,
  C      =   6.2240634,
  D      =   0.0502917,
  theta0 =  40.0,
  mH     =  -0.053583,
  bH     =   0.9995901,
  mW     =  -0.0021758,
  bW     =   0.1633299
)
```

The result above is **exactly 1** — and that is no coincidence. The SigLinDe
coefficients of every profile are calibrated so that the profile function equals
1 at the reference temperature of **8 °C**. At 8 °C the profile term drops out of
$Q(D) = KW \times h(\vartheta_D) \times F_{WT,D}$, so the customer value $KW$ is
simply the daily consumption on an 8 °C day — which is exactly how it is defined.

`HKO` is the one exception, and precisely because of its missing linear part.
The SigLinDe profiles owe that exact-1 value at 8 °C to the linear term, which
is fitted (together with the sigmoid) to pin $h(8\,\text{°C}) = 1$. `HKO` has
no linear part, so its bare sigmoid is not pinned to the reference and reaches
about **1.06** at 8 °C instead — visible in the BDEW profile sheet, where the
`HKO` row lists $h(8\,\text{°C}) = 1.05612$.

The temperature $\vartheta$ fed to this function is the daily *allocation temperature* (defined next); the dimensionless $h(\vartheta)$ is later scaled to kWh by the customer value $KW$ (covered below).

### Allocation temperature

The *allocation temperature* $\vartheta_D$ is the daily temperature fed to the
profile function $h(\vartheta)$. Two ways to compute it are defined in the
Leitfaden:[^allocation]

**Simple daily mean** — the arithmetic mean of the 24 hourly values over the
gas day:

$$\vartheta_D = \frac{1}{24} \sum_{i=1}^{24} T_i$$

**Geometrically-weighted 4-day mean** — recommended by BDEW for grid
operators, giving more weight to the current day:

$$\vartheta_D = \frac{T_D + 0.5 \cdot T_{D-1} + 0.25 \cdot T_{D-2} + 0.125 \cdot T_{D-3}}{1 + 0.5 + 0.25 + 0.125}$$

[^allocation]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2025-10-28. Berlin, section 3.5.2, p. 22.

`slp_gas()` accepts whichever values you pass in `temperatures` — the choice of
method is the caller's responsibility. If you have raw daily means from DWD and
want the geometrically-weighted variant, apply it before calling `slp_gas()`.
(If you already have gas forecast temperature (GPT) values from a meteorological
service provider, use those directly — GPT is not this formula.)

```{r allocation-temperature, eval=FALSE}
# Geometrically-weighted 4-day mean (more weight on the current day):
weights <- c(1, 0.5, 0.25, 0.125)

allocation_temp <- \(temps) {
  as.numeric(
    stats::filter(temps, weights / sum(weights),
                  method = "convolution", sides = 1)  # past values only
  )
}

# Apply to your daily mean series before passing it to slp_gas().
# The first three days come back NA (no 3-day history yet):
# temps_weighted <- allocation_temp(temps)
```

### Customer value (Kundenwert)

The customer value accounts for a customer’s individual consumption behaviour. It is a scaling factor in kWh/day — specifically, the daily gas consumption at the reference temperature of 8 °C, the point at which $h(\vartheta) = 1$ for every profile. We can rearrange the gas consumption equation above and write $KW$ as:

$$
KW = \frac{E_a}{\displaystyle\sum_{D} h(\vartheta_D) \times F_{WT,D}}
$$

Where 

* $E_a$ is the annual gas consumption
* $\vartheta_D$ is the daily allocation temperature for day $D$
* $F_{WT,D}$ is the weekday factor for day $D$

Using the example of a single-family home (`HEF`) in Düsseldorf with an annual
consumption of 15,000 kWh — the same customer as in the README — the table
below shows how the denominator is built up, day by day, from the Düsseldorf
long-term daily mean temperatures (DWD station 1078, 2004–2024):

```{r kundenwert_table, echo=FALSE, message=FALSE}
# Düsseldorf long-term daily mean temperature (DWD station 1078, mean over
# 2004-2024) -- the same reference series behind the Kundenwert in the README.
# Embedded here so the worked example is self-contained; see "Deriving the
# Kundenwert" below for how to fetch it.
ref_dates <- seq.Date(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day")
ref <- data.frame(
  date        = ref_dates,
  temp_c_mean = c(
    4.8, 4.2, 4.2, 4.6, 4.5, 4.0, 4.4, 4.6, 4.4, 4.2, 4.4, 4.8, 4.8,
    4.2, 3.6, 3.5, 3.5, 3.3, 3.1, 3.1, 2.5, 2.4, 2.2, 2.4, 2.1, 2.4,
    2.9, 3.3, 3.6, 3.4, 3.5, 4.1, 4.2, 4.6, 4.6, 4.7, 4.5, 3.1, 3.0,
    3.3, 3.4, 3.4, 2.9, 3.1, 4.2, 5.0, 5.6, 5.3, 4.9, 4.5, 5.1, 5.4,
    5.6, 5.2, 5.3, 5.0, 4.7, 3.9, 4.0, 4.6, 5.0, 4.9, 5.1, 4.5, 4.7,
    4.8, 5.7, 6.6, 6.6, 6.2, 6.3, 6.9, 6.8, 7.6, 8.3, 7.8, 7.2, 7.3,
    7.0, 7.0, 7.3, 7.2, 7.5, 7.7, 8.0, 8.3, 8.8, 8.8, 9.9, 9.9, 9.1,
    9.4, 9.6, 8.6, 8.1, 9.6, 10.2, 9.4, 9.7, 10.2, 10.0, 10.4, 10.2,
    10.6, 10.7, 10.9, 10.6, 10.9, 11.0, 11.3, 11.6, 11.5, 11.7, 12.3,
    12.3, 11.6, 11.5, 12.0, 12.3, 12.5, 12.5, 12.4, 12.1, 12.2, 12.2,
    13.1, 13.5, 14.1, 15.0, 14.9, 14.3, 13.7, 13.5, 13.5, 13.0, 13.3,
    13.1, 14.3, 14.5, 15.2, 16.0, 15.8, 15.1, 15.0, 15.3, 15.8, 16.0,
    16.2, 15.6, 15.4, 15.6, 15.9, 17.1, 17.1, 16.9, 16.7, 16.7, 17.0,
    17.6, 17.7, 17.9, 17.8, 16.9, 16.8, 17.2, 17.6, 17.0, 18.3, 19.2,
    18.4, 18.1, 17.6, 17.6, 18.3, 19.0, 19.1, 18.7, 19.0, 19.6, 19.4,
    19.2, 18.9, 19.2, 19.2, 19.5, 19.2, 18.7, 18.7, 18.6, 19.0, 19.1,
    18.9, 18.4, 18.4, 18.4, 19.2, 19.8, 20.0, 20.7, 21.1, 20.3, 19.8,
    20.1, 20.0, 20.1, 20.3, 20.4, 20.5, 20.0, 19.7, 19.6, 19.3, 20.1,
    19.7, 19.9, 19.6, 19.9, 20.3, 20.0, 18.8, 18.9, 18.5, 18.8, 19.1,
    19.0, 19.0, 19.3, 19.0, 18.6, 19.1, 19.3, 19.0, 18.7, 18.6, 19.0,
    19.1, 18.4, 18.5, 17.8, 17.9, 17.4, 17.2, 17.5, 16.8, 17.3, 17.5,
    17.7, 18.1, 17.9, 17.4, 17.2, 17.2, 17.3, 17.2, 16.6, 16.3, 15.8,
    15.8, 15.6, 15.0, 14.7, 14.4, 14.7, 15.0, 14.7, 14.8, 14.9, 14.1,
    14.1, 13.9, 13.9, 13.7, 13.4, 13.7, 13.4, 13.4, 13.3, 13.4, 13.3,
    13.2, 12.5, 12.1, 12.5, 12.0, 11.8, 12.3, 11.5, 11.3, 11.3, 11.5,
    11.4, 12.0, 11.9, 12.0, 11.3, 11.1, 11.9, 11.9, 12.1, 11.3, 10.9,
    11.0, 10.5, 10.3, 10.5, 10.7, 9.9, 9.7, 9.0, 8.9, 9.0, 8.8, 8.7,
    8.2, 8.0, 7.6, 8.0, 7.8, 7.9, 7.5, 7.3, 7.1, 6.8, 6.2, 6.0, 6.2,
    6.5, 6.0, 6.0, 5.2, 5.6, 4.9, 4.4, 4.0, 4.5, 4.3, 3.9, 4.1, 4.8,
    5.3, 5.6, 5.1, 4.5, 3.8, 4.0, 4.1, 3.5, 3.8, 4.4, 4.6, 4.7, 4.4,
    4.6, 3.9, 4.9, 6.3, 6.7, 6.5, 5.3, 4.5, 3.8, 3.7, 3.9, 4.7, 5.7
  )
)

# Compute h(theta) for HEF, variant 34
p <- standardlastprofile:::.gas_profile_params[["34"]][["HEF"]]
ref$h <- slp_gas_siglinde(
  ref$temp_c_mean, p$A, p$B, p$C, p$D, p$theta0,
  p$mH, p$bH, p$mW, p$bW
)

# Weekday factors (all 1 for HEF, but include for completeness)
wt_keys <- standardlastprofile:::.get_gas_weekday_key(ref$date)
fwt <- standardlastprofile:::.gas_weekday_factors[["HEF"]]
ref$f_wt <- unname(fwt[wt_keys])

# Show first and last few rows with ellipsis
show <- rbind(
  head(ref, 1),
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  ref[floor(nrow(ref) / 2) + c(-1, 0, 1), ],
  data.frame(date = NA, temp_c_mean = NA, h = NA, f_wt = NA),
  tail(ref, 1)
)
show_df <- data.frame(
  Date = ifelse(is.na(show$date), "\u22ee", format(show$date, "%b %e")),
  `°C` = ifelse(is.na(show$temp_c_mean), "\u22ee", round(show$temp_c_mean, 1)),
  `h(ϑ)` = ifelse(is.na(show$h), "\u22ee", round(show$h, 4)),
  check.names = FALSE
)
knitr::kable(show_df, align = "lrr")
```

The sum in the denominator across all `r nrow(ref)` days gives
$\sum_D h(\vartheta_D) \times F_{WT,D}$ = `r round(sum(ref$h * ref$f_wt), 2)`,
so $KW \approx$ 15,000 / `r round(sum(ref$h * ref$f_wt), 2)` =
**`r round(15000 / sum(ref$h * ref$f_wt), 1)` kWh/day**.

With $h(\vartheta_D)$ and $KW$ in hand, the only remaining factor in
$Q(D) = KW \times h(\vartheta_D) \times F_{WT,D}$ is the weekday factor $F_{WT,D}$.

### Weekday factors

Unlike the electricity profiles, which group Monday–Friday into a single
`workday` type, the gas procedure uses a separate scaling factor for every weekday 
from Monday through Sunday.[^fwt]

Public holidays are treated as a Sunday.[^holiday_rule]
24 December and 31 December are treated as Saturday unless they fall
on a Sunday; this convention mirrors the established rule from the BDEW
electricity SLP procedure and is applied here by analogy.

[^holiday_rule]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2025-10-28. Berlin, Annex 3, pp. 129–130.

[^fwt]: BDEW/VKU/GEODE (2025). *Leitfaden Abwicklung von Standardlastprofilen Gas*, as of 2025-10-28. Berlin, Annex 6, pp. 140–163 (per-profile data sheets).

For the residential profiles `HEF`, `HMF`, and `HKO`, however, all weekday factors
are equal to 1, that is, gas consumption in households is assumed not to differ
significantly by day of the week. 
Commercial profiles on the other hand show visible weekday patterns, the range of
the weekday factors is `r do.call(rbind, standardlastprofile:::.gas_weekday_factors) |> range() |> paste(collapse = " - ")`, illustrated by the chart below:

```{r weekday_factors_data, echo=FALSE}
days <- c("Mo", "Tu", "We", "Th", "Fr", "Sa", "Su")
profiles_f <- names(standardlastprofile:::.gas_weekday_factors)

fwt <- do.call(rbind, lapply(profiles_f, \(pid) {
  wfct <- standardlastprofile:::.gas_weekday_factors[[pid]]
  data.frame(
    profile_id = pid,
    day        = factor(names(wfct), levels = days),
    fct_value  = unname(wfct),
    category   = if(startsWith(pid, "H")) "household" else "commercial"
  )
}))
```

```{r weekday_factors_heatmap_static, echo=FALSE, fig.asp=1, fig.retina=2, eval=!knitr::is_html_output()}
#| fig.alt = "Heatmap of gas weekday factors with one row per profile and one
#|  column per day of the week, grouped into customer categories. Colour shows
#|  the percentage deviation from the weekly mean: commercial profiles are lower
#|  at weekends and higher on weekdays, while household profiles stay close to
#|  flat across all seven days."
ggplot(fwt, aes(day, profile_id, fill = fct_value - 1)) +
  geom_tile(width = .95, height = .95) +
  coord_equal(ratio = .75) +
  scale_fill_gradient2(NULL, labels = scales::label_percent(), midpoint = 0, low = "#311dfc", mid = "#fffafe", high = "#ff005e") +
  facet_grid(rows = vars(category), scales = "free", space = "free") +
  labs(
    title   = NULL,
    caption = NULL,
    x       = NULL,
    y       = NULL,
    fill    = NULL
  ) +
  theme_minimal() +
  theme(legend.position    = "bottom") +
  theme(legend.key.width = unit(0.099,"npc")) +
  theme(panel.grid = element_blank())
```

```{r weekday_factors_heatmap_html, echo=FALSE, results='asis', eval=knitr::is_html_output()}
fwt_json <- jsonlite::toJSON(
  data.frame(
    profile_id = fwt$profile_id,
    day        = as.character(fwt$day),
    fct_value  = fwt$fct_value,
    category   = fwt$category
  ),
  dataframe = "rows"
)
# Profile descriptions for y-axis tooltips
desc_pids <- unique(as.character(fwt$profile_id))
desc_map <- vapply(desc_pids, \(pid) {
  slp_info(pid)[[pid]]$description
}, character(1))
desc_json <- jsonlite::toJSON(as.list(desc_map), auto_unbox = TRUE)
cat(sprintf('
<div id="fwt-widget" style="max-width: 600px; margin: 1.5em auto; font-family: -apple-system, BlinkMacSystemFont, \'Segoe UI\', Roboto, sans-serif;">
<div id="fwt-chart"></div>
</div>
<script type="module">
import * as Plot from "https://cdn.jsdelivr.net/npm/@observablehq/plot@0.6/+esm";

const data = %s;
const descriptions = %s;
const days = ["Mo","Tu","We","Th","Fr","Sa","Su"];
const householdIds = [...new Set(data.filter(d => d.category === "household").map(d => d.profile_id))];
const commerceIds  = [...new Set(data.filter(d => d.category === "commercial").map(d => d.profile_id))];
const profileOrder = [...householdIds, ...commerceIds];

const chart = Plot.plot({
  marginLeft: 40,
  marginBottom: 30,
  height: profileOrder.length * 32 + 60,
  x: { domain: days, padding: 0.08, label: null },
  y: { domain: profileOrder, padding: 0.08, label: null },
  color: {
    type: "diverging",
    domain: [-0.6, 0.3],
    pivot: 0,
    range: ["#311dfc", "#fffafe", "#ff005e"],
    label: "Weekday factors for SLP gas profiles as deviation from 1",
    tickFormat: "+.0%%",
    legend: true
  },
  marks: [
    Plot.cell(data, {
      x: "day",
      y: "profile_id",
      fill: d => d.fct_value - 1,
      tip: true,
      title: d => {
        const v = d.fct_value - 1;
        return v === 0 ? "0%%" : (v > 0 ? "+" : "") + (v * 100).toFixed(1) + "%%";
      }
    }),
    Plot.ruleY([householdIds.length - 0.5], { stroke: "#999", strokeDasharray: "3,3" })
  ]
});

document.getElementById("fwt-chart").appendChild(chart);

// Add tooltips to y-axis tick labels
chart.querySelectorAll("text").forEach(el => {
  const pid = el.textContent.trim();
  if (!descriptions[pid]) return;
  el.style.cursor = "help";
  const t = document.createElementNS("http://www.w3.org/2000/svg", "title");
  t.textContent = descriptions[pid];
  el.prepend(t);
});
</script>
', fwt_json, desc_json))
```

## Profile IDs

There are 15 gas profile IDs defined in the BDEW Leitfaden, split into
residential and commercial / industrial types.

### Residential profiles

Residential profiles share two important characteristics: all weekday factors
equal 1 (no day-of-week differentiation), and heating load is the dominant
driver of demand variability.

| Profile | Description | Notes |
|---------|-------------|-------|
| `HEF`   | Single-family home (Einfamilienhaus) | Strongest seasonal swing; includes space heating + hot water |
| `HMF`   | Multi-family home (Mehrfamilienhaus) | Lower per-unit demand than `HEF` due to shared building envelope |
| `HKO`   | Cooking and hot water only (Kochen / Warmwasser) | No space heating; TUM profile — sigmoid only, linear part is zero |

`HKO` is appropriate for customers whose gas connection serves only a cooker
or instantaneous water heater. Its flat, temperature-insensitive shape is
visible in the SigLinDe curve plot above.

### Commercial and industrial profiles

Commercial profiles have profile-specific weekday factors that reflect the
operating hours and schedules of each sector. They are grouped here by
consumption pattern:

**Process-heat dominated** (relatively flat seasonal pattern, strong weekday
structure):

| Profile | Description |
|---------|-------------|
| `GWA`   | Laundries — large hot-water demand, insensitive to outdoor temperature |
| `GBA`   | Bakeries — continuous baking process, pronounced weekday factors |
| `GPD`   | Paper and printing — industrial process heat |

**Space-heating dominated** (strong winter peak, moderate weekday variation):

| Profile | Description |
|---------|-------------|
| `GKO`   | Small commercial (Kleinstgewerbe) |
| `GHA`   | Trade and commerce (Handel) |
| `GBD`   | Services (Dienstleistung) |
| `GBH`   | Accommodation (Beherbergung) |
| `GGA`   | Gastronomy (Gastronomie) |
| `GMK`   | Metal and automotive (Metall / Kfz) |
| `GGB`   | Mixed commercial (gemischtes Gewerbe) |
| `GMF`   | Large multi-family / mixed use (Mehrfamilienhaus groß) |

**Aggregate / summary profile**:

| Profile | Description |
|---------|-------------|
| `GHD`   | Trade, commerce and services aggregate (GHD-Stützpunkt) — weighted mean across all G-types |

`GHD` is used when a customer cannot be assigned to a specific sector. Its
parameters are a weighted average of the individual commercial profiles and
its weekday factors are close to but not equal to 1.

## A working example: Düsseldorf

With the algorithm covered, let us see how to use `slp_gas()` in practice.
`slp_gas()` takes:
- a `profile_id`
- a `dates` vector, and a matching `temperatures` vector, and 
- a `kundenwert`.

`kundenwert` is the customer-specific scaling factor in kWh/day. 

### Deriving the Kundenwert

To produce a meaningful result the temperature series should cover a full
year at least. You can use measured data from a trustworthy
institution like the German Meteorological Service, or DWD for short. The
[`rdwd`](https://cran.r-project.org/package=rdwd) package provides a
straightforward interface to the open-data portal of the DWD 
(Deutscher Wetterdienst).[^dwd] No registration or API key is required.

[^dwd]: DWD Open Data: <https://opendata.dwd.de/>

#### rdwd package 

The workflow has three steps:

1. `selectDWD()` — find the download URL for a station and dataset.
2. `dataDWD()` — download and unzip the file.
3. `readDWD()` — parse the file into a data frame.

For gas SLPs, we need the **daily climate summary** (`var = "kl"`,
`res = "daily"`). The column `TMK` contains the daily mean temperature in °C.
To find the exact station name to pass to `selectDWD()`, use `findID()` or open
the interactive station map:

```{r dwd_stations, eval=FALSE}
rdwd::findID("Freiburg")             # returns matching station names and IDs
rdwd::selectDWD(mapDWD = TRUE)       # opens an interactive map in the browser
```

The Kundenwert is derived once, from the customer's annual consumption and a
**reference temperature series for their location**, using
`slp_gas_kundenwert()`. The resulting value is then passed to `slp_gas()` for
whatever period you want to generate — a single month, a quarter, the whole
year.

Keeping the derivation a separate step is deliberate. $KW$ is a property of the
customer and their local climate, so it should be computed from a
representative full year — ideally a multi-year mean. 

Take the single-family home (`HEF`) from the README's example: located in
**Düsseldorf**, with an annual gas consumption of **15,000 kWh**. For the
reference temperatures we use the **long-term daily mean over 2004–2024** — a
multi-year (climatological) mean rather than a single year, so that no
individual-year anomaly distorts the scaling factor^[BDEW Leitfaden, section
3.6.3]. 

```{r kw_derive, eval=FALSE}
library(rdwd)

# Full daily climate record for Düsseldorf (DWD station 1078)
links <- selectDWD("Duesseldorf", res = "daily", var = "kl",
                   per = c("historical", "recent"))
raw   <- do.call(rbind, readDWD(dataDWD(links, read = FALSE, force = TRUE),
                                varnames = FALSE))

# Long-term mean daily temperature over 2004-2024, one value per calendar day
ref       <- raw[format(raw$MESS_DATUM, "%Y") %in% as.character(2004:2024), ]
ref$mmdd  <- format(ref$MESS_DATUM, "%m-%d")
clim      <- aggregate(TMK ~ mmdd, data = ref, FUN = mean)
clim      <- clim[clim$mmdd != "02-29", ]          # drop the leap day
dates_ref <- as.Date(paste0("2023-", clim$mmdd))   # any non-leap year
temps_ref <- clim$TMK

# kundenwert for a 15,000 kWh/a HEF customer
slp_gas_kundenwert("HEF", dates_ref, temps_ref, annual_consumption = 15000)
#>      HEF
#> 55.08344
```

This Düsseldorf customer has a Kundenwert of about **55.1 kWh/day**. That fixed
value is what the README example passes to `slp_gas()` to generate the
2025/26 heating season — and it is what we use below to ask how the same
customer would fare in a different climate.

### What if the customer moves?

We keep the Düsseldorf customer's Kundenwert of **55.1 kWh/day** fixed and ask
what they would consume in a colder climate — same customer, same KW, different
weather. Germany spans several distinct climate zones; we compare three:

* **Chemnitz** — continental climate in the Erzgebirge foothills of Saxony;
  colder winters and the greatest temperature amplitude of the three.
* **Freiburg im Breisgau** — mild oceanic-continental climate in the sheltered
  Upper Rhine Plain; the warmest, sunniest large city in Germany.
* **Hamburg** — maritime climate on the North Sea coast; mild but cloudy,
  windy winters.

The grid below has one column per comparison city and one row per month of the
2025/26 heating season (the same period as the README example). Each point is a
day, plotting the city's daily gas consumption (y-axis) against Düsseldorf
(x-axis); the 45° line marks equal consumption.

That winter all three cities ran colder than Düsseldorf, so every point cloud
sits above the line — the question is by how much:

- **Chemnitz** is highest, about **35 % more gas** over the season
  (≈ 16,400 vs 12,200 kWh);
- **Hamburg** follows at about **25 % more**;
- **Freiburg** is closest to Düsseldorf at about **11 % more**.

The gap is widest on the cold January and February days at the top-right of each
panel, and narrows in the mild shoulder months of October and April where the
clouds approach the line.

```{r gas_cities_chart, echo=FALSE, fig.asp=2.4, fig.retina=2, message=FALSE}
#| fig.alt = "Faceted scatterplot grid: columns are Chemnitz, Freiburg, Hamburg;
#|  rows are months October to April. Each point is a day; the x-axis is daily
#|  gas consumption in Düsseldorf, the y-axis in the comparison city, with a
#|  45-degree reference line. All three cities' point clouds sit above the line
#|  that winter, furthest for Chemnitz (about 35 percent more), then Hamburg
#|  (about 25 percent), with Freiburg closest to the line (about 11 percent)."
standardlastprofile:::.slp_plot_gas_cities()
```
